---
title: "Alzheimer’s Disease Progression Under Dropout: Longitudinal Modelling of BPRS and Severe CDR-SB"
author: 
  - Moses Mburu
date: "`r format(Sys.Date(), '%B %d, %Y')`"
fontsize: 8pt 
output:
  beamer_presentation:
    incremental: false
    latex_engine: xelatex
  slidy_presentation:
    incremental: false
  github_document:
    df_print: kable
  ioslides_presentation:
    widescreen: true
    css: style.css
    self_contained: true
    incremental: false
header-includes:
- \usepackage{longtable}
- \usepackage{tabu}
- \usepackage{xcolor}
- \usepackage{bookmark}
- \usepackage{booktabs}
- \usepackage{xcolor}
- \setbeamercolor{item}{fg=black}
- \setbeamercolor{subitem}{fg=black}
- \setbeamercolor{subsubitem}{fg=black}
- \usetheme{Hannover} #- \usetheme{Boadilla}
- \usepackage{lmodern}          # Latin Modern (clean, scalable)
- \usepackage{microtype}        # Improves spacing & kerning subtly
- \setbeamertemplate{sidebar left}{}
- \setbeamertemplate{navigation symbols}{}

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, 
                      warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.5)

library(here)
library(tidyverse)
library(data.table)
library(haven)
library(janitor)
library(knitr)
library(nlme)
library(gridExtra)
library(kableExtra)
# source(here("R","lda_homework1_2025.R"))
# save.image(here("R","LDA1.RData"))
load(here("hm3","hm3_objects.rda"))
```


---

### Study Description

- Study design: Longitudinal Alzheimer cohort with repeated follow-up visits per patient (baseline + yearly visits).
- Outcomes:
  - Brief Psychiatric Rating Scale (BPRS) (continuous psychiatric symptom score)
  - Clinical Dementia Rating – Sum of Boxes (CDR-SB) severe indicator:
    \(Z_{ij}=\mathbb{I}(\text{CDR-SB}_{ij}>10)\)
- Research questions
  - BPRS trajectory: Does BPRS change over time, and is its evolution associated with baseline characteristics (e.g., age/sex/education, baseline severity) and biomarkers (time-varying and/or baseline; e.g., ABPET/TAUPET)?
  - Severe cognition risk: Does the probability of severe cognition (\(\text{CDR-SB}>10\)) increase over time, and is that increase associated with baseline variables and biomarkers?
- Challenge:
  - follow-up is incomplete due to dropout, with missingness increasing over time, so naïve complete-case analysis can be biased.

---

### Methodology/Analysis of Data with missing data approaches

- Complete-case analysis (CC): analyse only subjects with complete follow-up (or only fully observed outcome vectors).
  - Simple, but can be biased unless missingness is MCAR (missing completely at random).

- last observation carried forward (LOCF): use all observed visits, and for dropout replace missing post-dropout values with the last observed value.
  - Uses more data than CC, but LOCF imposes a strong assumption that outcomes remain unchanged after dropout and can be biased if dropout is related to disease status.

- Multiple imputation (MI):

  - Impute missing outcomes multiple times from an imputation model and combine estimates.
  - Typically relies on MAR given the imputation model and included history/covariates.

- Likelihood-based models under MAR (primary approach):

  - Fit a longitudinal model (e.g., LMM/GLMM) by maximum likelihood (or REML for LMM).
  
- Weighted GEE (WGEE):

  - Extend GEE to handle dropout by weighting observed data by the inverse probability of being observed.
- Sensitivity analysis for MNAR:

  - Because MAR cannot be verified from the observed data alone, evaluate robustness under plausible departures using a pattern-mixture / \(\delta\)-shift approach.

---

### Results

```{r, fig.cap = "Distribution of CDRSB Scores & BPRS Over Time", fig.width=10, fig.height=4.5}


grid.arrange(cdrsb_time_plot,
    simple_avg_plot,
    ncol = 2 )
```

The left panel shows that, as follow-up progresses, a growing proportion of patients move into the worse CDR-SB category (CDR-SB (>10)), indicating worsening cognition over time. The right panel shows a clear increasing LOESS trend in BPRS across years, suggesting psychiatric symptom burden also increases with time before we interpret these trends formally, we next examine missingness and dropout patterns.

---

#### Missing Data Patterns

```{r, fig.width=10, fig.height=4.5}

grid.arrange(boxplot_meanbprso0,
             missing_by_time_plot,
             ncol = 2)
```

* Missingness increases over time, reaching about 60% at the last follow-up.
* Participants with more missed visits have higher baseline BPRS, suggesting dropout is related to observed severity (not MCAR).


---

```{r}
   
top_patterns2 <- top_patterns |>
  dplyr::mutate(
    pattern = gsub("\\s+", "", pattern_string),   # remove spaces/newlines
    pct = round(N / nrow(alzheimer_data) * 100, 1)
  ) |>
  dplyr::select(
    `Missingness type` = missingness_type,
    `Pattern (O=obs, M=miss)` = pattern,
    `n` = N,
    `%` = pct
  )

top_patterns2 <- top_patterns2 |>
  dplyr::mutate(
    n_obs = stringr::str_count(`Pattern (O=obs, M=miss)`, "O")
  ) |>
  dplyr::arrange(dplyr::desc(n_obs), `Pattern (O=obs, M=miss)`) |>
  dplyr::select(-n_obs)


top_patterns2 |>
  kable(
    format = "latex",
    booktabs = TRUE,
    longtable = FALSE,
    align = "lccc",
    escape = TRUE,
    caption = "Most common missingness patterns (O = observed, M = missing)"
  ) |>
  kableExtra::kable_styling(
    latex_options = c( "scale_down"),
    font_size = 9,
    position = "center"
  ) |>
  kableExtra::row_spec(0, bold = TRUE) |>
  kableExtra::column_spec(2, width = "2.6cm")


```


* The most common patterns show monotone dropout (once missing starts, later visits are usually missing).


---


### Key findings (BPRS LMM + severe cognition models)

BPRS (LMM)

- BPRS increased  over follow-up: +7.220 points/year (95% CI 7.141–7.299), p < 0.001
- Baseline factors associated with higher BPRS levels:

  - Older age (p < 0.001), higher BMI (p < 0.001), nursing home residence (p < 0.001)
  - Employment (paid job) was associated with lower BPRS (p < 0.001)
  - Sex was borderline (female vs male: p = 0.067)
  - ADL and baseline CDR-SB were not significant main effects
- Baseline severity moderated progression:
  - time × baseline CDR-SB = −0.022, p < 0.001
  - Interpretation: worse baseline CDR-SB → slower increase in BPRS over time (convergence/ceiling effect)

Severe cognition (CDR-SB > 10; GLMM)

- Risk increased  with time: OR = 1.986 per year (95% CI 1.783–2.211), p < 0.001
- Biomarker modification:
  - time × tau PET OR = 1.473, p = 0.020  higher tau accelerates progression to severe status
- Other covariates were not significant in this subject-specific model:
  - sex, age, BMI, ADL, amyloid PET, and tau PET main effect (all p > 0.05)

---


### Conclusion (what the clinician needs to hear)

-  Models estimate how BPRS and severe cognition (CDR-SB > 10) change over time and which patient factors are associated with faster worsening, helping identify higher-risk patients for closer follow-up.
- Sensitivty Analysis\(\delta\)-shift? When we assumed post-dropout  (\(\delta=0\) to \(3\)) shift, the main conclusions did not change: the time trend remained strongly positive and the key predictors (age, BMI, nursing home, paid job, and the time × baseline CDR-SB effect) stayed consistent, while weaker effects (sex, ADL, baseline CDR-SB main effect) were not robust.

