---
title: "Longitudinal Analysis of BPRS  in Alzheimer's Disease"
author: 
  - Anita Kerubo Ogero
  - Moses Mburu
date: "`r format(Sys.Date(), '%B %d, %Y')`"
fontsize: 8pt 
output:
  beamer_presentation:
    incremental: false
    latex_engine: xelatex
  slidy_presentation:
    incremental: false
  github_document:
    df_print: kable
  ioslides_presentation:
    widescreen: true
    css: style.css
    self_contained: true
    incremental: false
header-includes:
- \usepackage{longtable}
- \usepackage{tabu}
- \usepackage{xcolor}
- \usepackage{bookmark}
- \usepackage{booktabs}
- \usepackage{xcolor}
- \setbeamercolor{item}{fg=black}
- \setbeamercolor{subitem}{fg=black}
- \setbeamercolor{subsubitem}{fg=black}
- \usetheme{Hannover} #- \usetheme{Boadilla}
- \usepackage{lmodern}          # Latin Modern (clean, scalable)
- \usepackage{microtype}        # Improves spacing & kerning subtly
- \setbeamertemplate{sidebar left}{}
- \setbeamertemplate{navigation symbols}{}

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, 
                      warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.5)

library(here)
library(tidyverse)
library(data.table)
library(haven)
library(janitor)
library(knitr)
library(nlme)
library(gridExtra)
library(kableExtra)
# source(here("R","lda_homework1_2025.R"))
# save.image(here("R","LDA1.RData"))
load(here("R","LDA1.RData"))
```


---

### Study description

- **Population:** 1,253 patients with Alzheimer’s disease enrolled in **25 clinical trials**.  
- **Outcome:** Brief Psychiatric Rating Scale (BPRS), measured annually from baseline to year 6  
  - Up to **7 measurements per patient** (baseline + 6 follow-up years)  
  - Higher BPRS = more severe psychiatric symptoms  
- **Design:** Unbalanced longitudinal data  
  - Not all patients are observed at all time points  
  - Total number of BPRS observations: **6,220**  
- **Baseline covariates:**
  - Demographic: sex, age, education  
  - Clinical: BMI, ADL score, living situation (home vs nursing home), job status (paid vs no paid job)  
  - Disease severity / biomarkers: baseline CDR-SB, amyloid PET (ABPET0), tau PET (TAUPET0)

---

#### Missing data analysis

```{r graph_1, fig.cap = "Missingness Analysis: Baseline BPRS by Number of Missed Visits (left) and Percentage Missing Over Time (right)",  fig.width = 8.5, fig.height = 4.5}

grid.arrange(boxplot_meanbprso0,
             missing_by_time_plot,
             ncol = 2)

```

---

#### BPRS over time

```{r simple_avg_plots, fig.cap = "Average BPRS Scores Over Time", fig.width = 6.5, fig.height = 3}

simple_avg_plot

```



### Individual profiles

```{r spaghetti_plots, fig.cap = "Spaghetti Plots of BPRS Over Time by Clinical Center", fig.width = 5.5, fig.height =2.9}

spaggheti_plot

```

- These 50 individual profiles show that patients start at very different BPRS levels but almost all worsen over time, with broadly similar upward slopes.

---

#### Separate Linear Models for Each Follow-up Year

```{r}
kable(summary_dt,
  format = "latex",
  booktabs = TRUE,
  caption = "Separate Linear Models for Each Follow-up Year",
  align = "lrrrrrrr", # l=left, r=right alignment for columns
  linesep = "",
  font.size = 9
) |>
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

- The biomarker effects (amyloid and tau PET) jump around from year to year and their CIs are often wide or include zero.
- The other variables shows more consistent effects across years.

---

#### Conclusions and modeling implications



* Dropout is higher in patients with worse baseline BPRS.
* For population averages we use a repeated-measures model with an unstructured covariance; for patient-specific effects we use a mixed model with patient random intercept and slope plus a trial random intercept.

---

### Two-Stage Model Results


```{r, fig.cap = "Distribution of Individual Intercepts and Slopes from two stage analysis", fig.width = 6, fig.height = 2.8}
two_stage
```

* Starting levels vary a lot Intercepts are widely spread (20–100 BPRS), showing big differences in baseline severity between patients.
* Progression rates are similar Slopes are tightly clustered around  6- 7 points/year, so most patients worsen at a similar rate.
* Bias Patients with higher BPRS are more likely to drop out, the slopes mainly reflect those who stay in follow-up.

---

```{r}
tidy_with_ci(slope_model) |>
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Two-Stage Model Results for Individual Slopes",
    align = "lrrrr", # l=left, r=right alignment for columns
    linesep = ""
  ) |>
  column_spec(2, width = "5cm") |> # Set the width of the 2nd column (estimate)
  kable_styling(latex_options = c("hold_position", "scale_down"))
```



* Only two baseline factors show an association with the rate of change patients living in a nursing home worsen slightly faster (about $0.2$ points more per year), while those with higher baseline CDR-SB show a slightly slower increase in BPRS (about $0.02$ points less per year per CDR-SB point).

---

### Mixed Model with Random Intercept and Slope Results

```{r}

# summary(lmer_reduced_mean_structure)
lmer_tidy <- tidy_with_ci(lmer_reduced_mean_structure)

lmer_tidy |>
  kable(
    format = "latex",
    caption = "Mixed Model with Random Interceptand Slope Results",
    booktabs = TRUE,
    longtable =FALSE,
    align = "lrrrr"
  ) |>
  column_spec(2, width = "5cm") |>
  kable_styling(font_size = 8, position = "center") |>
  kable_styling(latex_options = c("hold_position", "scale_down"))

```

* On average, BPRS increases by about $\approx 7.2$ points per year (time effect), confirming steady worsening over follow-up.
* Higher age, higher BMI, living in a nursing home, and not having a paid job are all associated with higher BPRS levels at any given time.
* The interaction $\text{time} \times \text{cdrsb0}$ is negative: patients with worse baseline cognition (higher CDR-SB) show a slightly slower yearly increase in BPRS.


---


#### Diagnostics Plots

```{r diagnostics_plots, fig.cap = "Mixed Model Residual Diagnostics", fig.width = 8, fig.height = 3}
grid.arrange(
    p1_diag_q5, p2_diag_q5,
    ncol = 2
)

```

* Residuals are scattered evenly around zero with no clear pattern, so the linear fit and constant variance look reasonable.
* In the Q–Q plot, points lie close to the straight line, so the normality assumption for residuals is roughly satisfied (only small deviations in the tails).


---

#### Comparing with Two Stage Results

```{r compare_two_stage_mixed, fig.cap = "Comparison of Two-Stage and Mixed Model Estimates", fig.width = 8, fig.height = 3}

gridExtra::grid.arrange(
  p_int,
  p_slope,
  ncol = 2
)

```

* Intercepts from the two-stage and mixed models are highly correlated ( r = 0.94), so both approaches give almost the same starting BPRS level.
* Slopes  correlate well but less strongly (r = 0.77); the mixed model pulls extreme slopes towards the average.

---

### Recommendations for future similar studies

Based on what we observed in the exploratory plots and models, we suggest the following design and data collection improvements for future work on BPRS in Alzheimer patients.

1. Address informative dropout and missingness: Correlation of missingness and baseline BPRS: Enroll comparable patients(severity bracket)
2. Standardize clinical site procedures High variability between sites: Also introduce a patient subgroup to indicate severity

---
