---
title: "Longitudinal Analysis of BPRS  in Alzheimer's Disease"
author: 
  - Anita Kerubo Ogero
  - Moses Mburu
date: "`r format(Sys.Date(), '%B %d, %Y')`"
fontsize: 8pt 
output:
  beamer_presentation:
    incremental: false
    latex_engine: xelatex
  slidy_presentation:
    incremental: false
  github_document:
    df_print: kable
  ioslides_presentation:
    widescreen: true
    css: style.css
    self_contained: true
    incremental: false
header-includes:
- \usepackage{longtable}
- \usepackage{tabu}
- \usepackage{xcolor}
- \usepackage{bookmark}
- \usepackage{booktabs}
- \usepackage{xcolor}
- \setbeamercolor{item}{fg=black}
- \setbeamercolor{subitem}{fg=black}
- \setbeamercolor{subsubitem}{fg=black}
- \usetheme{Hannover} #- \usetheme{Boadilla}
- \usepackage{lmodern}          # Latin Modern (clean, scalable)
- \usepackage{microtype}        # Improves spacing & kerning subtly
- \setbeamertemplate{sidebar left}{}
- \setbeamertemplate{navigation symbols}{}

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE, 
                      warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.5)

library(here)
library(tidyverse)
library(data.table)
library(haven)
library(janitor)
library(knitr)
library(nlme)
library(gridExtra)
library(kableExtra)
# source(here("R","lda_homework1_2025.R"))
# save.image(here("R","LDA1.RData"))
load(here("R","LDA2.RData"))
```


---

### Study description

- **Population:** 1,253 patients with Alzheimer’s disease enrolled in **25 clinical trials**.  
- **Outcome:** Brief Psychiatric Rating Scale (BPRS), measured annually from baseline to year 6  
  - Up to **7 measurements per patient** (baseline + 6 follow-up years)  
  - Higher BPRS = more severe psychiatric symptoms  
- **Design:** Unbalanced longitudinal data  
  - Not all patients are observed at all time points  
  - Total number of BPRS observations: **6,220**  
- **Baseline covariates:**
  - Demographic: sex, age, education  
  - Clinical: BMI, ADL score, living situation (home vs nursing home), job status (paid vs no paid job)  
  - Disease severity / biomarkers: baseline CDR-SB, amyloid PET (ABPET0), tau PET (TAUPET0)

---



```{r}
tidy_with_ci(glmm_center, exponentiate = TRUE) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    # longtable = TRUE,
    align = "lcccccc"
  ) |>
  # scale down to fit page width
  kableExtra::kable_styling(
    latex_options = c(
      "scale_down"
    ),
    font_size = 9
  )
```



- what you see 
- interpretation of coefficients


---

```{r, fig.width=10, fig.height=4.5, fig.cap="Predicted BPRS Trajectories over Time by Baseline CDR-SB"}

grid.arrange(p_int, p_slope, ncol = 2)

```


---


