---
title: "LDA Homework 2 (2025-2026)"
author: "\\vspace{1cm}\n\\ \nAnita Kerubo Ogero (2469491) \\\nStudent B (student number)
  \\\nStudent C (student number) \\\nStudent D (student number)\n\\textbf{\\large{Group
  members:}} \\\n"
output:
  html_document:
    toc: true
    df_print: paged
  pdf_document:
    toc: true
    pdf_engine: xelatex
  github_document: default
always_allow_html: true
fontsize: 11pt
header-includes:
- \renewcommand{\baselinestretch}{1.1}
- \usepackage{graphicx}
- \usepackage{xcolor}         % replaces 'color'
- \usepackage{amsmath,amssymb}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}       % optional
- \usepackage{makecell}       % optional
- \usepackage{pdflscape}
- \usepackage{float}          % for [H] (hold_position)
- \usepackage{fancyhdr}       % if customizing header/footer
---





```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                        message = FALSE,
                        warning = FALSE,
                        fig.width = 7,
                        fig.height = 4.3)

library(kableExtra)

use_latex_packages()


#list.files(lda_data_path)

#' Tidy model with 95% CIs and a formatted estimate column
#'
#' Works for lm/glm via broom and lme4/glmmTMB via broom.mixed.
#' @param model a fitted model (lm, glm, lmerMod, glmerMod, glmmTMB, etc.)
#' @param effects which effects to return
#' for mixed models ("fixed", "ran_pars", "ran_vals")
#' @param conf_level CI level
#' @param exponentiate exponentiate estimates/CIs (useful for logit/Poisson)
#' @param digits rounding for formatted strings
#' @return tibble with columns term, estimate (formatted),
#' std.error, statistic, p.value, conf.low, conf.high
tidy_with_ci <- function(model,
                         effects = c("fixed"),
                         conf_level = 0.95,
                         exponentiate = FALSE,
                         digits = 2) {
  mm_class <- class(model)

  is_mixed <-
    any(c(
      "lmerMod", "lmerModLmerTest",
      "glmerMod", "nlmerMod"
    ) %in% mm_class) ||
      inherits(model, "glmmTMB")

  if (is_mixed) {
    if (!requireNamespace("broom.mixed", quietly = TRUE)) {
      stop("broom.mixed is required for mixed models.\n
        Install with install.packages('broom.mixed').")
    }
    out <- broom.mixed::tidy(
      model,
      effects = effects,
      conf.int = TRUE,
      conf.level = conf_level,
      exponentiate = exponentiate
    )
    # keep only common columns
    out <- out[, intersect(
      c(
        "effect", "group", "term", "estimate", "std.error",
        "statistic", "p.value", "conf.low", "conf.high"
      ),
      names(out)
    )]
  } else if (inherits(model, "gls")) {
    # nlme::gls objects are not supported by broom::tidy; extract manually
    s <- summary(model)
    if (!is.null(s$tTable)) {
      tt <- as.data.frame(s$tTable)
      # tTable columns: Value, Std.Error, t-value, p-value
      out <- data.frame(
        term = rownames(tt),
        estimate = tt[["Value"]],
        std.error = tt[["Std.Error"]],
        statistic = tt[["t-value"]],
        p.value = tt[["p-value"]],
        stringsAsFactors = FALSE
      )
      # Wald (normal) CIs
      z <- stats::qnorm(1 - (1 - conf_level) / 2)
      out$conf.low <- out$estimate - z * out$std.error
      out$conf.high <- out$estimate + z * out$std.error
      # ensure column names match later expectations
      out <- out[, c(
        "term", "estimate", "std.error",
        "statistic", "p.value", "conf.low", "conf.high"
      )]
    } else {
      stop("Unable to extract coefficient table from gls summary output.")
    }
  } else {
    if (!requireNamespace("broom", quietly = TRUE)) {
      stop("broom is required. Install with install.packages('broom').")
    }
    out <- broom::tidy(
      model,
      conf.int = TRUE,
      conf.level = conf_level,
      exponentiate = exponentiate
    )
    out <- out[, intersect(
      c(
        "term", "estimate", "std.error",
        "statistic", "p.value", "conf.low", "conf.high"
      ),
      names(out)
    )]
  }

  # format numeric columns
  num_cols <- intersect(c("estimate", "std.error", "statistic", "
  p.value", "conf.low", "conf.high"), names(out))
  out[num_cols] <- lapply(out[num_cols], function(x) {
    if (is.numeric(x)) round(x, digits) else x
  })

  # formatted estimate with CI
  if (all(c("estimate", "conf.low", "conf.high") %in%
    names(out))) {
    # Use sprintf to format numbers to a consistent width for alignment
    # make 3f to be digitsf
    out$estimate_str <- sprintf(paste0("%7.", digits, "f"), out$estimate)
    out$conf.low_str <- sprintf(paste0("%7.", digits, "f"), out$conf.low)
    out$conf.high_str <- sprintf(paste0("%7.", digits, "f"), out$conf.high)

    est_ci <- paste0(
      out$estimate_str, " [",
      out$conf.low_str, ", ", out$conf.high_str, "]"
    )
    out$estimate <- est_ci
  }

  # optional: pretty p-values (e.g., <0.001)
  if ("p.value" %in% names(out)) {
    out$p.value <- ifelse(is.na(out$p.value), NA_character_,
      ifelse(out$p.value < 10^(-digits),
        paste0("<", format(10^(-digits), scientific = FALSE)),
        format(round(out$p.value, digits), nsmall = digits)
      )
    )
  }

  # reorder columns: put term first, then estimate string
  keep_order <- c(
    "effect", "group", "term", "estimate", "std.error",
    "statistic", "p.value"
  )
  out <- out[, intersect(keep_order, names(out))]

  tibble::as_tibble(out)
}


# Tidy adjacent model comparisons from base anova() output
# Works for anova(m1, m2, ...): lm/glm, lmer/glmer, gls/lme, etc.
tidy_anova_compare <- function(anova_df, digits = 2) {
  x <- anova_df
  is_num <- vapply(x, is.numeric, logical(1))
  x[is_num] <- lapply(x[is_num], function(v) {
    ifelse(is.na(v), v, round(v, digits))
  })
  x$call <- NULL
  x
}

```



```{r}
library(here)
library(tidyverse)
library(data.table)
library(haven)
library(janitor)
library(knitr)
library(lme4)   
library(lmerTest)
library(pbkrtest)
library(gridExtra)

lda_path <- here()
lda_data_path <- here(lda_path, "data")
#list.files(lda_data_path)

```



```{r}
# Load the data
alzheimer_data2 <- read_sas(here(
    lda_data_path,
    "alzheimer25.sas7bdat"
)) |> clean_names()


```





```{r factor-conversion}
setDT(alzheimer_data2)
# log bmi
#alzheimer_data[, bmi := log(bmi)]
factor_info <- list(
  sex = list(
    levels = c(0, 1),
    labels = c("Male", "Female")
  ),
  edu = list(
    levels = c(1, 2, 3, 4),
    labels = c("Primary", "Lower secondary", "Upper secondary", "Higher")
  ),
  job = list(
    levels = c(0, 1),
    labels = c("No Paid Job", "Paid Job")
  ),
  wzc = list(
    levels = c(0, 1),
    labels = c("Lives at Home", "Nursing Home")
  )
)


for (col_name in names(factor_info)) {
  info <- factor_info[[col_name]]
  alzheimer_data2[, (col_name) := factor(get(col_name),
    levels = info$levels,
    labels = info$labels
  )]
}

# Verify the changes by checking the structure of the new columns
#kable(alzheimer_data[1:5, c("sex", "edu", "job", "wzc")])
```





```{r reshape-data}

id_vars <- c("patid", "trial", "sex", "age", 
             "edu", "bmi", "inkomen", "job", "adl", "wzc")
      



alzheimer_long2 <- melt(alzheimer_data2,
  id.vars = id_vars,
  measure.vars = patterns(
    cdrsb = "^cdrsb[0-6]",
    bprs = "^bprs[0-6]",
    abpet = "^abpet[0-6]",
    taupet = "^taupet[0-6]"
  ),
  variable.name = "time",
  value.name = c("cdrsb", "bprs", "abpet", "taupet"),
  variable.factor = FALSE
)

alzheimer_long2[, time := as.integer(time)]

#alzheimer_long[, time := time - 1]


# Display the first few rows of the new long-format data
#kable(head(alzheimer_long))
```


```{r}
names(alzheimer_long2)
```

# Question 1

step 1 : Create the dichotomized outcome (CDRSBbin)

```{r}
alzheimer_long2 <- alzheimer_long2 |>
  mutate(
    cdrsb_bin = ifelse(cdrsb <= 10, 0, 1)
  )

```


```{r}
#View(alzheimer_long2)
```

step 2 : Adjust for sex, age, BMI (baseline confounders)

```{r}
base_model <- glm(
  cdrsb_bin ~ sex + age + bmi,
  data = alzheimer_long2,
  family = binomial
)

summary(base_model)

```

Step 3 — Add ADL, ABPET, TAUPET to the model

```{r}
full_model <- glm(
  cdrsb_bin ~ sex + age + bmi + adl + abpet + taupet,
  data = alzheimer_long2,
  family = binomial
)

summary(full_model)

```

```{r}
AIC(base_model, full_model)

```

Here, AIC is the preferred criterion when the goal is prediction or evaluating whether additional variables improve model fit. BIC penalizes model complexity more heavily and is more appropriate when the goal is selecting the “true” model or when sample size is very large.We are checking whether ADL, ABPET, and TAUPET add information beyond sex, age, and BMI. That is a predictive/model improvement question, so AIC is the correct choice.


## Summary of Results
We first fit a logistic regression adjusting only for sex, age, and BMI. In this base model, female sex (p = 0.0126) and age (p < 0.0001) were significantly associated with lower odds of severe CDRSB status, while BMI showed no association.

Next, we extended the model to include ADL, ABPET, and TAUPET. In the full model, ADL remained non-significant (p = 0.486), whereas both ABPET (p = 0.0046) and TAUPET (p < 0.0001) were strong predictors of severe impairment.

Model comparison using AIC indicated improved fit for the full model (AIC = 8086) compared to the base model (AIC = 8134). This shows that biomarker information (ABPET, TAUPET) provides additional predictive value beyond demographic factors.


## Discussion

After adjusting for sex, age, and body mass index, we examined whether daily functioning (ADL) and PET biomarkers predicted severe cognitive impairment. The results show that PET markers for amyloid and tau are significant and meaningful predictors of worse CDRSB status. Patients with higher amyloid or tau burden are more likely to fall into the more severely impaired category. Measures of daily functioning and BMI were not helpful predictors. Adding the biomarker information clearly improved the predictive accuracy of the model.



# Question 2

```{r}
library(dplyr)
library(ggplot2)

# Make sure cdrsb_bin and time exist and have no weird values
# str(alzheimer_long)

max_time <- max(alzheimer_long2$time, na.rm = TRUE)
lags <- 1:max_time

semivar_df <- lapply(lags, function(h) {
  tmp <- alzheimer_long2 %>%
    group_by(patid) %>%
    arrange(time, .by_group = TRUE) %>%
    mutate(
      y_t  = cdrsb_bin,
      y_th = dplyr::lead(cdrsb_bin, n = h),
      diff2 = (y_t - y_th)^2
    ) %>%
    filter(!is.na(diff2))  # keep only valid pairs at lag h
  
  data.frame(
    lag = h,
    semivariance = 0.5 * mean(tmp$diff2)
  )
}) %>%
  bind_rows()

# Plot semivariogram
ggplot(semivar_df, aes(x = lag, y = semivariance)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Empirical Semivariogram of CDRSBbin Over Time",
    x = "Time lag (years)",
    y = "Semivariance"
  ) +
  theme_minimal()
```


## Formulate and fit a marginal model to the longitudinal binary profiles.



We model the longitudinal binary outcome \(Y_{it} = \mathrm{CDRSBbin}_{it}\) for patient \(i\) at visit \(t\).

The marginal mean is

\[
\pi_{it} = \mathbb{E}(Y_{it} \mid X_{it})
= \Pr(Y_{it} = 1 \mid X_{it}),
\]
with logit link
\[
\log\left(\frac{\pi_{it}}{1 - \pi_{it}}\right)
=
\beta_0 
+ \beta_1 \,\text{time}_{it}
+ \beta_2 \,\text{sex}_i
+ \beta_3 \,\text{age}_i
+ \beta_4 \,\text{bmi}_i
+ \beta_5 \,\text{adl}_{it}
+ \beta_6 \,\text{abpet}_{it}
+ \beta_7 \,\text{taupet}_{it}.
\]

Because \(Y_{it}\) is binary, the marginal variance is specified by the Bernoulli model:
\[
\mathrm{Var}(Y_{it} \mid X_{it}) 
= \pi_{it}\,(1 - \pi_{it}).
\]
For subject \(i\), the marginal covariance matrix is written in the usual GEE form
\[
\mathrm{Var}(\mathbf{Y}_i \mid X_i)
= A_i^{1/2}\, R(\alpha)\, A_i^{1/2},
\]
where \(A_i\) is a diagonal matrix with entries
\(\pi_{it}(1-\pi_{it})\), and \(R(\alpha)\) is the working correlation matrix.  
In our analysis we use an independence working correlation, i.e.
\(R(\alpha) = I\), motivated by the weak empirical within-subject correlation and many subjects observed only once.

### Motivation for Model Choice

Generalized Estimating Equations (GEE) were used because the aim of the analysis is to estimate marginal, population-averaged effects of covariates on the probability of having CDRSBbin = 1 over time. The outcome is binary and repeatedly measured within individuals, which induces within-subject correlation. GEE appropriately accounts for this correlation and provides consistent coefficient estimates even if the working correlation structure is misspecified. Since the scientific interest is in population level effects—not subject-specific trajectories—GEE is more appropriate than a mixed-effects logistic model. Additionally, GEE is robust to missing data under a MAR mechanism, which is relevant given the observed dropout patterns.


### Working Correlation Matrix


Exploratory analysis showed that although CDRSB was measured longitudinally, most subjects had very limited within-subject variation, with many never changing state. The semivariogram indicated that correlation declines with lag, but GEE correlation estimates for all structures (exchangeable, AR1) were essentially zero, suggesting negligible residual correlation after covariate adjustment. Exchangeable was inappropriate because correlation was not constant across lags; AR(1) was theoretically possible but unsupported by the data. Therefore, the independence working correlation was selected because it provides stable estimates and is consistent with the near-zero empirical correlation observed in the data.




```{r}
library(geepack)

gee_cdrsb_ar1 <- geeglm(
  cdrsb_bin ~ time + sex + age + bmi + adl + abpet + taupet,
  id      = patid,
  data    = alzheimer_long2,
  family  = binomial(link = "logit"),
  corstr  = "independence"
)

summary(gee_cdrsb_ar1)

```

Exploratory transition analysis showed that 31% of patients never changed binary CDRSB status, and 44% changed only once. This indicates very little within-subject variation over time. As a result, the empirical semivariogram showed increasing variability with time lag, but the GEE correlation parameter (α) could not be estimated under AR(1) or exchangeable structures and defaulted to zero. Therefore, an independence working correlation structure is appropriate and supported by the data.


```{r}
library(dplyr)

transitions <- alzheimer_long2 %>%
  arrange(patid, time) %>%
  group_by(patid) %>%
  summarise(
    n_obs = sum(!is.na(cdrsb_bin)),
    transitions = sum(cdrsb_bin[-1] != cdrsb_bin[-n()], na.rm = TRUE)
  )



```


```{r}
library(ggplot2)

ggplot(transitions, aes(x = transitions)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Number of Transitions (0↔1) per Patient",
    x = "Transitions",
    y = "Number of Patients"
  ) +
  theme_minimal()

```



```{r}
prop.table(table(transitions$transitions))

```

### Results


A marginal logistic regression model was fitted using GEE with an independence working correlation structure to examine factors associated with worsening cognitive impairment (CDRSBbin = 1) over time. The analysis included 6,220 observations from repeated annual measurements.

The results showed that time was strongly associated with increased odds of worse cognitive impairment (estimate = 0.25, p < 0.001). None of the baseline covariates reached strong statistical significance at the 5% level, but several showed borderline associations. Female sex showed slightly lower odds (estimate = –0.10, p = 0.095), older age showed a weak negative association (estimate = –0.013, p = 0.077), and tau PET showed a positive trend (estimate = 0.33, p = 0.052). ADL, BMI, and amyloid PET were not significant in this model.

### Discussion

Using a population-based model that accounts for repeated measures, we found that patients’ cognitive impairment (as measured by CDR-SB categories) worsens steadily over time. None of the baseline clinical or biomarker variables showed strong independent effects after adjusting for sex, age, and BMI, though tau PET showed a borderline trend toward higher risk. Functional status (ADL), amyloid PET, and BMI did not meaningfully predict change in this dichotomized measure. Overall, the progression of cognitive impairment over time was the dominant signal in the data rather than static baseline characteristics.




# Question 3

## Random-effects model for longitudinal binary CDRSB profiles

### Random‐Effects Model for Longitudinal Binary CDRSB Profiles

We model the binary outcome \( \text{CDRSBbin}_{it} \) for patient \( i \) at time \( t \)
using a logistic mixed‐effects model with patient‐specific random intercepts and slopes,
and a trial‐specific random intercept.

The marginal mean is

\[
\pi_{it} = \Pr(\text{CDRSBbin}_{it} = 1 \mid X_{it}, b_{0i}, b_{1i}, u_{0j}),
\]

with the logit link:

\[
\log\!\left( \frac{\pi_{it}}{1 - \pi_{it}} \right)
=
\beta_0
+ \beta_1\,\text{time}_{it}
+ \beta_2\,\text{sex}_i
+ \beta_3\,\text{age}_i
+ \beta_4\,\text{bmi}_i
+ \beta_5\,\text{adl}_{it}
+ \beta_6\,\text{abpet}_{it}
+ \beta_7\,\text{taupet}_{it}
+ b_{0i}
+ b_{1i}\,\text{time}_{it}
+ u_{0j},
\]

where

- \( b_{0i} \) = patient-specific random intercept  
- \( b_{1i} \) = patient-specific random slope for time  
- \( u_{0j} \) = trial-level random intercept  

The random effects are assumed to follow:

\[
\begin{pmatrix}
b_{0i} \\
b_{1i}
\end{pmatrix}
\sim
N\!\left(
\begin{pmatrix} 0 \\ 0 \end{pmatrix},
\;
\begin{pmatrix}
\sigma_{0}^2 & \sigma_{01} \\
\sigma_{01} & \sigma_{1}^2
\end{pmatrix}
\right),
\qquad
u_{0j} \sim N(0, \tau^2).
\]




```{r}
library(lme4)

glmer_cdrsb <- glmer(
  cdrsb_bin ~ time + sex + age + bmi + adl + abpet + taupet +
    1 + (time | patid) + (1 | trial),         # patient + centre random intercepts
  data   = alzheimer_long2,
  family = binomial
)

summary(glmer_cdrsb)
```


```{r}
library(broom.mixed)

fixef_table <- tidy(glmer_cdrsb, effects = "fixed")  # estimates, SE, z, p
rand_sd      <- as.data.frame(VarCorr(glmer_cdrsb))  # variance components

```




## Empirical Bayes predictions of the random effects


The “empirical Bayes predictions” are the patient-specific random intercepts  b0i and (and trial random intercepts u0j



```{r}
eb <- ranef(glmer_cdrsb)$patid
eb$patid <- rownames(eb)
colnames(eb) <- c("intercept_EB", "patid")

library(ggplot2)

ggplot(eb, aes(intercept_EB)) +
  geom_histogram(bins = 40, fill = "skyblue", color = "black") +
  labs(
    title = "Empirical Bayes Predictions of Random Intercepts",
    x = "EB Intercept",
    y = "Frequency"
  ) +
  theme_minimal()


```




```{r}
re <- ranef(glmer_cdrsb)

patient_re <- as.data.frame(re$patid)
patient_re$patid <- rownames(re$patid)

colnames(patient_re) <- c("b0i_intercept", "b1i_slope", "patid")

trial_re <- as.data.frame(re$trial)
trial_re$trial <- rownames(re$trial)
colnames(trial_re) <- c("u0j_intercept", "trial")

```




```{r}
ggplot(patient_re, aes(b0i_intercept)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  labs(title = "Empirical Bayes: Patient Random Intercepts",
       x = "b0i", y = "Frequency") +
  theme_minimal()

```

The histogram of patient-specific random intercepts shows large variability in baseline probability of having severe CDRSB impairment.
Some patients start much lower (large negative intercepts), while others start much higher (large positive intercepts).
This indicates substantial heterogeneity in baseline disease severity that is not explained by sex, age, BMI, ADL, ABPET, or TAUPET.



```{r}
ggplot(patient_re, aes(b1i_slope)) +
  geom_histogram(bins = 30, fill = "lightgreen", color = "black") +
  labs(title = "Empirical Bayes: Patient Random Slopes",
       x = "b1i", y = "Frequency") +
  theme_minimal()

```
The random slope histogram also shows large variation.
Some patients experience rapid worsening over time (positive slopes), whereas others show slow or even improving trajectories (negative slopes).
This means progression speed differs strongly across individuals, even after adjusting for covariates.

```{r}
ggplot(patient_re, aes(b0i_intercept, b1i_slope)) +
  geom_point(alpha = 0.4) +
  theme_minimal() +
  labs(title = "Random Intercept vs Random Slope",
       x = "Intercept (b0i)",
       y = "Slope (b1i)")

```

The scatterplot of random intercept vs. random slope shows a very strong negative correlation.
Patients who start with worse impairment (high intercepts) tend to decline more slowly, while patients with mild baseline symptoms tend to deteriorate faster.



## Discussion


Patient-specific random effects showed large differences in both the baseline severity of cognitive impairment and the rate of decline over time. Importantly, individuals who started with worse impairment tended to deteriorate more slowly, while those with milder symptoms showed steeper progression. These findings highlight substantial biological variability across patients that is not captured by standard clinical predictors.

