---
title: "LDA Homework 2 (2025-2026)"
author: 
  - Anita Kerubo Ogero (2469491)  
  - Okolie Tess Eucharia (2468507)  
  - Moses Mburu (2469245)  
  - Dorothy Chepkoech (2469284)

format:
  pdf:
    pdf-engine: xelatex
    fig-pos: "H"
    tbl-pos: "H"
    number-depth: 6
    include-in-header:
      text: |
        \renewcommand{\baselinestretch}{1.1}
        \usepackage{graphicx}
        \usepackage{xcolor}
        \usepackage{amsmath,amssymb}
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{array}
        \usepackage{multirow}
        \usepackage{makecell}
        \usepackage{pdflscape}
        \usepackage{float}
        \usepackage{fancyhdr}
  gfm: default

fontsize: 11pt
---





```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                        message = FALSE,
                        warning = FALSE,
                        fig.width = 7,
                        fig.height = 4.3)

library(kableExtra)

use_latex_packages()


#list.files(lda_data_path)

#' Tidy model with 95% CIs and a formatted estimate column
#'
#' Works for lm/glm via broom and lme4/glmmTMB via broom.mixed.
#' @param model a fitted model (lm, glm, lmerMod, glmerMod, glmmTMB, etc.)
#' @param effects which effects to return
#' for mixed models ("fixed", "ran_pars", "ran_vals")
#' @param conf_level CI level
#' @param exponentiate exponentiate estimates/CIs (useful for logit/Poisson)
#' @param digits rounding for formatted strings
#' @return tibble with columns term, estimate (formatted),
#' std.error, statistic, p.value, conf.low, conf.high
tidy_with_ci <- function(model,
                         effects = c("fixed"),
                         conf_level = 0.95,
                         exponentiate = FALSE,
                         digits = 2) {
  mm_class <- class(model)

  is_mixed <-
    any(c(
      "lmerMod", "lmerModLmerTest",
      "glmerMod", "nlmerMod"
    ) %in% mm_class) ||
      inherits(model, "glmmTMB")

  if (is_mixed) {
    if (!requireNamespace("broom.mixed", quietly = TRUE)) {
      stop("broom.mixed is required for mixed models.\n
        Install with install.packages('broom.mixed').")
    }
    out <- broom.mixed::tidy(
      model,
      effects = effects,
      conf.int = TRUE,
      conf.level = conf_level,
      exponentiate = exponentiate
    )
    # keep only common columns
    out <- out[, intersect(
      c(
        "effect", "group", "term", "estimate", "std.error",
        "statistic", "p.value", "conf.low", "conf.high"
      ),
      names(out)
    )]
  } else if (inherits(model, "gls")) {
    # nlme::gls objects are not supported by broom::tidy; extract manually
    s <- summary(model)
    if (!is.null(s$tTable)) {
      tt <- as.data.frame(s$tTable)
      # tTable columns: Value, Std.Error, t-value, p-value
      out <- data.frame(
        term = rownames(tt),
        estimate = tt[["Value"]],
        std.error = tt[["Std.Error"]],
        statistic = tt[["t-value"]],
        p.value = tt[["p-value"]],
        stringsAsFactors = FALSE
      )
      # Wald (normal) CIs
      z <- stats::qnorm(1 - (1 - conf_level) / 2)
      out$conf.low <- out$estimate - z * out$std.error
      out$conf.high <- out$estimate + z * out$std.error
      # ensure column names match later expectations
      out <- out[, c(
        "term", "estimate", "std.error",
        "statistic", "p.value", "conf.low", "conf.high"
      )]
    } else {
      stop("Unable to extract coefficient table from gls summary output.")
    }
  } else {
    if (!requireNamespace("broom", quietly = TRUE)) {
      stop("broom is required. Install with install.packages('broom').")
    }
    out <- broom::tidy(
      model,
      conf.int = TRUE,
      conf.level = conf_level,
      exponentiate = exponentiate
    )
    out <- out[, intersect(
      c(
        "term", "estimate", "std.error",
        "statistic", "p.value", "conf.low", "conf.high"
      ),
      names(out)
    )]
  }

  # format numeric columns
  num_cols <- intersect(c("estimate", "std.error", "statistic", "
  p.value", "conf.low", "conf.high"), names(out))
  out[num_cols] <- lapply(out[num_cols], function(x) {
    if (is.numeric(x)) round(x, digits) else x
  })

  # formatted estimate with CI
  if (all(c("estimate", "conf.low", "conf.high") %in%
    names(out))) {
    # Use sprintf to format numbers to a consistent width for alignment
    # make 3f to be digitsf
    out$estimate_str <- sprintf(paste0("%7.", digits, "f"), out$estimate)
    out$conf.low_str <- sprintf(paste0("%7.", digits, "f"), out$conf.low)
    out$conf.high_str <- sprintf(paste0("%7.", digits, "f"), out$conf.high)

    est_ci <- paste0(
      out$estimate_str, " [",
      out$conf.low_str, ", ", out$conf.high_str, "]"
    )
    out$estimate <- est_ci
  }

  # optional: pretty p-values (e.g., <0.001)
  if ("p.value" %in% names(out)) {
    out$p.value <- ifelse(is.na(out$p.value), NA_character_,
      ifelse(out$p.value < 10^(-digits),
        paste0("<", format(10^(-digits), scientific = FALSE)),
        format(round(out$p.value, digits), nsmall = digits)
      )
    )
  }

  # reorder columns: put term first, then estimate string
  keep_order <- c(
    "effect", "group", "term", "estimate", "std.error",
    "statistic", "p.value"
  )
  out <- out[, intersect(keep_order, names(out))]

  tibble::as_tibble(out)
}


# Tidy adjacent model comparisons from base anova() output
# Works for anova(m1, m2, ...): lm/glm, lmer/glmer, gls/lme, etc.
tidy_anova_compare <- function(anova_df, digits = 2) {
  x <- anova_df
  is_num <- vapply(x, is.numeric, logical(1))
  x[is_num] <- lapply(x[is_num], function(v) {
    ifelse(is.na(v), v, round(v, digits))
  })
  x$call <- NULL
  x
}

```



```{r}
library(here)
library(tidyverse)
library(data.table)
library(haven)
library(janitor)
library(knitr)
library(lme4)   
library(lmerTest)
library(pbkrtest)
library(gridExtra)

lda_path <- here()
lda_data_path <- here(lda_path, "data")
#list.files(lda_data_path)

```



```{r}
# Load the data
alzheimer_data <- read_sas(here(
    lda_data_path,
    "alzheimer25.sas7bdat"
)) |> clean_names()


```





```{r factor-conversion}
setDT(alzheimer_data)
# log bmi
#alzheimer_data[, bmi := log(bmi)]
factor_info <- list(
  sex = list(
    levels = c(0, 1),
    labels = c("Male", "Female")
  ),
  edu = list(
    levels = c(1, 2, 3, 4),
    labels = c("Primary", "Lower secondary", "Upper secondary", "Higher")
  ),
  job = list(
    levels = c(0, 1),
    labels = c("No Paid Job", "Paid Job")
  ),
  wzc = list(
    levels = c(0, 1),
    labels = c("Lives at Home", "Nursing Home")
  )
)


for (col_name in names(factor_info)) {
  info <- factor_info[[col_name]]
  alzheimer_data[, (col_name) := factor(get(col_name),
    levels = info$levels,
    labels = info$labels
  )]
}

# Verify the changes by checking the structure of the new columns
#kable(alzheimer_data[1:5, c("sex", "edu", "job", "wzc")])
```





```{r reshape-data}

id_vars <- c("patid", "trial", "sex", "age", 
             "edu", "bmi", "inkomen", "job", "adl", "wzc")
             #"cdrsb0",  "abpet0", "taupet0")



alzheimer_long <- melt(alzheimer_data,
  id.vars = id_vars,
  measure.vars = patterns(
    cdrsb = "^cdrsb[0-6]",
    bprs = "^bprs[0-6]",
    abpet = "^abpet[0-6]",
    taupet = "^taupet[0-6]"
  ),
  variable.name = "time",
  value.name = c("cdrsb", "bprs", "abpet", "taupet"),
  variable.factor = FALSE
)

alzheimer_long[, time := as.integer(time)]

alzheimer_long[, CDRSBbin := ifelse(cdrsb > 10, 1, 0)]
#alzheimer_long[, time := time - 1]


# Display the first few rows of the new long-format data
#kable(head(alzheimer_long))
```





### Question 1: Marginal Model (GEE)

Fit a marginal model to the binary response `CDRSBbin`.
Correct for: `sex`, `age`, `bmi`.
Examine effect of: `adl`, `abpet`, `taupet`.
Since `abpet` and `taupet` are time-varying, we include them as such. `adl` is baseline.

```{r}
#| label: cdrsb-stacked-plots
#| fig.height: 10
#| fig.width: 10
#| fig.cap: "Distribution of CDRSB Scores by Time and Site"

library(ggplot2)
library(scales)

# 1. Stacked Bar Plot: CDRSB by Time
# Calculate counts and percentages within each time point
plot_data_time <- alzheimer_long[!is.na(cdrsb), .N, by = .(time, CDRSBbin)]
plot_data_time[, pct := N / sum(N), by = time]

cdrsb_time_plot <- ggplot(plot_data_time, aes(
    x = factor(time),
    y = pct, fill = factor(CDRSBbin)
)) +
    geom_bar(stat = "identity", position = "fill") +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_brewer(palette = "Dark2") +
    labs(
        title = "Distribution of CDRSB Scores Over Time",
        x = "Time (Visit)",
        y = "Percentage",
        fill = "CDRSB Score"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")

# 2. Stacked Bar Plot: CDRSB by Site (trial)
# Calculate counts and percentages within each trial
plot_data_site <- alzheimer_long[!is.na(cdrsb), .N, 
by = .(trial, CDRSBbin)]
plot_data_site[, pct := N / sum(N), by = trial]

cdrsb_site_plot <- ggplot(plot_data_site, aes(
    x = factor(trial),
    y = pct, fill = factor(CDRSBbin)
)) +
    geom_bar(stat = "identity", position = "fill") +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_brewer(palette = "Dark2") +
    labs(
        title = "Distribution of CDRSB Scores by Site",
        x = "Site (Trial)",
        y = "Percentage",
        fill = "CDRSB Score"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90))

grid.arrange(cdrsb_time_plot, 
cdrsb_site_plot, ncol = 1)
```





```{r}
#| label: gee-model
#| tbl-cap: "GEE Model Results (Odds Ratios)"

library(geepack)
# Ensure data is sorted for GEE
setorder(alzheimer_long, patid, time)

gee_model <- geeglm(
    CDRSBbin ~ time + sex + age + bmi + adl + abpet + taupet,
    id = patid,
    data = alzheimer_long,
    family = binomial(link = "logit"),
    corstr = "exchangeable",
    std.err = "san.se" # Robust standard errors
)

summary(gee_model)

# Tidy output with Odds Ratios
tidy_with_ci(gee_model, exponentiate = TRUE) %>%
    kable(
        format = "latex",
        booktabs = TRUE,
        # longtable = TRUE,
        align = "lcccccc",
    ) |>
    # scale down to fit page width
    kableExtra::kable_styling(
        latex_options = c(
            "scale_down"
        ),
        font_size = 9
    )
```

**Interpretation:**
(Add interpretation here based on output)

# Question 2: Random-Effects Model (GLMM)

Fit a random-effects model (GLMM). Given the differences between sites observed previously, we include a random intercept for `trial` in addition to the random intercept for `patid`.

```{r glmm-model}
#| label: glmm-model
#| tbl-cap: "GLMM Model Results (Odds Ratios)"


glmm_model <- glmer(
  CDRSBbin ~ time + sex + age + bmi + adl + abpet + taupet + (1 | trial) + (1 | patid),
  data = alzheimer_long,
  family = binomial(link = "logit"),
  nAGQ = 1 # Laplace approximation
)

summary(glmm_model)

tidy_with_ci(glmm_model, exponentiate = TRUE) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    # longtable = TRUE,
    align = "lcccccc"
  ) |>
  # scale down to fit page width
  kableExtra::kable_styling(
    latex_options = c(
      "scale_down"
    ),
    font_size = 9
  )
```

## Empirical Bayes Predictions

Examine the empirical Bayes predictions (random effects) to identify subjects with specific profiles.

```{r eb-predictions}
#| label: eb-predictions
#| fig-cap: "Empirical Bayes Predictions of Random Effects"
#| fig-width: 10
#| fig-height: 5
#| 
# Extract random effects
r_effs <- ranef(glmm_model)

# 1. Patient-level random effects
ranef_pat <- as.data.frame(r_effs$patid) %>%
  rownames_to_column("patid") %>%
  rename(random_intercept = `(Intercept)`)

# 2. Trial-level random effects
ranef_trial <- as.data.frame(r_effs$trial) %>%
  rownames_to_column("trial") %>%
  rename(random_intercept = `(Intercept)`)

# Plot distribution of Patient Random Intercepts
patid_plot <- ggplot(ranef_pat, aes(x = random_intercept)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Patient Random Intercepts", 
       subtitle = "Variation between subjects",
       x = "Log-Odds", y = "Count")

# Plot Trial Random Intercepts
trial_plot <- ggplot(ranef_trial, aes(x = reorder(trial, random_intercept), y = random_intercept)) +
  geom_point(size = 3, color = "darkred") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Trial Random Intercepts", 
       subtitle = "Variation between sites",
       x = "Trial", y = "Log-Odds Deviation")

grid.arrange(patid_plot, trial_plot, ncol = 2)


```


```{r}
#| label: extreme-subjects
#| tbl-cap: "Subjects with Highest Random Intercepts (High Risk)"
top_risk <- ranef_pat %>% arrange(desc(random_intercept))

head(top_risk, 10) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    # longtable = TRUE,
    align = "lcccccc"
  ) |>
  # scale down to fit page width
  kableExtra::kable_styling(
    latex_options = c(
      "scale_down"
    ),
    font_size = 9
  )


```


```{r}
#| label: extreme-subjects-low
#| tbl-cap: "Subjects with Lowest Random Intercepts (Low Risk)"
low_risk <- ranef_pat %>% arrange(random_intercept)
head(low_risk, 10) %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    # longtable = TRUE,
    align = "lcccccc"
  ) |>
  # scale down to fit page width
  kableExtra::kable_styling(
    latex_options = c(
      "scale_down"
    ),
    font_size = 9
  )

```
