---
title: "LDA Homework 3 (2025-2026)"
author: 
  - Anita Kerubo Ogero (2469491)  
  - Okolie Tess Eucharia (2468507)  
  - Moses Mburu (2469245)  
  - Dorothy Chepkoech (2469284)

format:
  pdf:
    pdf-engine: xelatex
    fig-pos: "H"
    tbl-pos: "H"
    number-depth: 6
    include-in-header:
      text: |
        \renewcommand{\baselinestretch}{1.1}
        \usepackage{graphicx}
        \usepackage{xcolor}
        \usepackage{amsmath,amssymb}
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{array}
        \usepackage{multirow}
        \usepackage{makecell}
        \usepackage{pdflscape}
        \usepackage{float}
        \usepackage{fancyhdr}
  gfm: default

fontsize: 11pt
---





```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                        message = FALSE,
                        warning = FALSE,
                        fig.width = 7,
                        fig.height = 4.3)

library(kableExtra)

use_latex_packages()


tidy_with_ci <- function(model,
                         effects = c("fixed", "ran_pars"),
                         conf_level = 0.95,
                         exponentiate = FALSE,
                         digits = 2) {
  mm_class <- class(model)

  is_mixed <-
    any(c(
      "lmerMod", "lmerModLmerTest",
      "glmerMod", "nlmerMod"
    ) %in% mm_class) ||
      inherits(model, "glmmTMB")

  if (is_mixed) {
    if (!requireNamespace("broom.mixed", quietly = TRUE)) {
      stop("broom.mixed is required for mixed models.\n
        Install with install.packages('broom.mixed').")
    }
    out <- broom.mixed::tidy(
      model,
      effects = effects,
      conf.int = TRUE,
      conf.level = conf_level,
      exponentiate = exponentiate
    )
    # keep only common columns
    out <- out[, intersect(
      c(
        "effect", "group", "term", "estimate", "std.error",
        "statistic", "p.value", "conf.low", "conf.high"
      ),
      names(out)
    )]
  } else if (inherits(model, "gls")) {
    # nlme::gls objects are not supported by broom::tidy; extract manually
    s <- summary(model)
    if (!is.null(s$tTable)) {
      tt <- as.data.frame(s$tTable)
      # tTable columns: Value, Std.Error, t-value, p-value
      out <- data.frame(
        term = rownames(tt),
        estimate = tt[["Value"]],
        std.error = tt[["Std.Error"]],
        statistic = tt[["t-value"]],
        p.value = tt[["p-value"]],
        stringsAsFactors = FALSE
      )
      # Wald (normal) CIs
      z <- stats::qnorm(1 - (1 - conf_level) / 2)
      out$conf.low <- out$estimate - z * out$std.error
      out$conf.high <- out$estimate + z * out$std.error
      # ensure column names match later expectations
      out <- out[, c(
        "term", "estimate", "std.error",
        "statistic", "p.value", "conf.low", "conf.high"
      )]
    } else {
      stop("Unable to extract coefficient table from gls summary output.")
    }
  } else {
    if (!requireNamespace("broom", quietly = TRUE)) {
      stop("broom is required. Install with install.packages('broom').")
    }
    out <- broom::tidy(
      model,
      conf.int = TRUE,
      conf.level = conf_level,
      exponentiate = exponentiate
    )
    out <- out[, intersect(
      c(
        "term", "estimate", "std.error",
        "statistic", "p.value", "conf.low", "conf.high"
      ),
      names(out)
    )]
  }

  # formatted estimate with CI
  if (all(c("estimate", "conf.low", "conf.high") %in% names(out))) {
    # Identify rows that have valid CIs (i.e., not ran_pars)
    has_ci <- !is.na(out$conf.low)
    
    # Create the formatted estimate string, applying CI only where valid
    out$estimate <- ifelse(
      has_ci,
      paste0(
        sprintf(paste0("%.", digits, "f"), out$estimate), " [",
        sprintf(paste0("%.", digits, "f"), out$conf.low), ", ",
        sprintf(paste0("%.", digits, "f"), out$conf.high), "]"
      ),
      sprintf(paste0("%.", digits, "f"), out$estimate) # Just format the estimate for ran_pars
    )
  }

  # format numeric columns that are not the main estimate
  num_cols <- intersect(c("std.error", "statistic"), names(out))
  out[num_cols] <- lapply(out[num_cols], function(x) {
    if (is.numeric(x)) round(x, digits) else x
  })

  # optional: pretty p-values (e.g., <0.001)
  if ("p.value" %in% names(out)) {
    # Format p-values, leaving NAs as they are
    out$p.value <- ifelse(is.na(out$p.value), NA_character_,
      ifelse(out$p.value < 10^(-digits),
        paste0("<0.", paste(rep("0", digits - 1), collapse = ""), "1"),
        format(round(out$p.value, digits), nsmall = digits)
      )
    )
  }

  # reorder columns: put term first, then estimate string
  keep_order <- c(
    "effect", "group", "term", "estimate", "std.error",
    "statistic", "p.value"
  )
  out <- out[, intersect(keep_order, names(out))]

  tibble::as_tibble(out)
}



# Tidy adjacent model comparisons from base anova() output
# Works for anova(m1, m2, ...): lm/glm, lmer/glmer, gls/lme, etc.
tidy_anova_compare <- function(anova_df, digits = 2) {
  x <- anova_df
  is_num <- vapply(x, is.numeric, logical(1))
  x[is_num] <- lapply(x[is_num], function(v) {
    ifelse(is.na(v), v, round(v, digits))
  })
  x$call <- NULL
  x
}



display_random_effects <- function(
    model,
    caption = "Random effects: variance components",
    digits = 3) {
  df <- tidy_re_varcorr(model, digits = digits)
  knitr::kable(
    df,
    format = "latex",
    booktabs = TRUE,
    #caption = caption,
    align = c("l", "l", "r", "r", "r")
  ) |>
    kableExtra::kable_styling(latex_options = "hold_position")
}



tidy_re_varcorr <- function(model, digits = 3) {
  vc <- lme4::VarCorr(model)
  rows <- list()

  for (grp in names(vc)) {
    m <- vc[[grp]]
    sd <- attr(m, "stddev")
    trm <- names(sd)
    var <- sd^2

    corMat <- tryCatch(attr(m, "correlation"),
      error = function(e) NULL
    )

    for (k in seq_along(trm)) {
      term <- trm[k]
      corr_val <- NA_real_
      if (!is.null(corMat) && nrow(corMat) > 1 &&
        term != "(Intercept)" &&
        "(Intercept)" %in% colnames(corMat)) {
        corr_val <- as.numeric(corMat[term, "(Intercept)"])
      }
      rows[[length(rows) + 1L]] <- data.frame(
        Groups = grp,
        Name = term,
        Variance = var[k],
        Std.Dev. = sd[k],
        Corr = corr_val,
        check.names = FALSE
      )
    }
  }

  out <- do.call(rbind, rows)

  # Add Residual variance
  resid_sd <- sigma(model)
  resid_var <- resid_sd^2
  resid_row <- data.frame(
    Groups = "Residual",
    Name = "",
    Variance = resid_var,
    Std.Dev. = resid_sd,
    Corr = NA_real_,
    check.names = FALSE
  )
  out <- rbind(out, resid_row)

  num_cols <- c("Variance", "Std.Dev.", "Corr")
  out[num_cols] <- lapply(
    out[num_cols],
    function(x) {
      ifelse(is.na(x),
        NA, round(x, digits)
      )
    }
  )
  out
}


```



```{r}
library(here)
library(tidyverse)
library(data.table)
library(haven)
library(janitor)
library(knitr)
library(lme4)   
library(lmerTest)
library(pbkrtest)
library(gridExtra)

lda_path <- here()
lda_data_path <- here(lda_path, "data")
#list.files(lda_data_path)

```



```{r}
# Load the data
alzheimer_data <- read_sas(here(
    lda_data_path,
    "alzheimer25.sas7bdat"
)) |> clean_names()


```





```{r factor-conversion}
setDT(alzheimer_data)
# log bmi
#alzheimer_data[, bmi := log(bmi)]
factor_info <- list(
  sex = list(
    levels = c(0, 1),
    labels = c("Male", "Female")
  ),
  edu = list(
    levels = c(1, 2, 3, 4),
    labels = c("Primary", "Lower secondary", "Upper secondary", "Higher")
  ),
  job = list(
    levels = c(0, 1),
    labels = c("No Paid Job", "Paid Job")
  ),
  wzc = list(
    levels = c(0, 1),
    labels = c("Lives at Home", "Nursing Home")
  )
)


for (col_name in names(factor_info)) {
  info <- factor_info[[col_name]]
  alzheimer_data[, (col_name) := factor(get(col_name),
    levels = info$levels,
    labels = info$labels
  )]
}

# Verify the changes by checking the structure of the new columns
#kable(alzheimer_data[1:5, c("sex", "edu", "job", "wzc")])
```


```{r reshape-data}

id_vars <- c("patid", "trial", "sex", "age", 
             "edu", "bmi", "inkomen", "job", "adl", "wzc",
            "cdrsb0",  "abpet0", "taupet0" )



alzheimer_long <- melt(alzheimer_data,
  id.vars = id_vars,
  measure.vars = patterns(
    cdrsb = "^cdrsb[0-6]",
    bprs = "^bprs[0-6]",
    abpet = "^abpet[0-6]",
    taupet = "^taupet[0-6]"
  ),
  variable.name = "time",
  value.name = c("cdrsb", "bprs", "abpet", "taupet"),
  variable.factor = FALSE
)

alzheimer_long[, time := as.integer(time)]

#alzheimer_long[, time := time - 1]


# Display the first few rows of the new long-format data
#kable(head(alzheimer_long))
```

# Abstract

# Introduction

Alzheimer’s disease is a progressive neurodegenerative disorder characterised by a gradual decline in cognitive, psychiatric, and functional abilities. Longitudinal studies play a central role in understanding disease progression and identifying patient characteristics associated with changes in clinical outcomes over time. However, the analysis of longitudinal data is often complicated by within-subject correlation, unbalanced follow-up, and missing data due to dropout, particularly in studies involving elderly or clinically vulnerable populations (Diggle et al., 2002; Fitzmaurice et al., 2011).

The present study follows a cohort of patients diagnosed with Alzheimer’s disease who were recruited across multiple clinical centres and assessed annually for up to six years. Psychiatric symptom severity was measured repeatedly using the Brief Psychiatric Rating Scale (BPRS), alongside cognitive and biomarker assessments. Baseline demographic and clinical characteristics—including age, sex, education, body mass index, functional status, employment status, and living situation—were recorded at study entry. As expected in long-term follow-up studies of Alzheimer’s disease, not all participants remained under observation for the full duration of the study, resulting in unbalanced longitudinal data with increasing missingness over time.

The primary objective of this analysis is to characterise the longitudinal evolution of BPRS and to investigate how psychiatric symptom trajectories relate to baseline patient characteristics and disease severity. To address these aims, a sequence of modelling approaches is adopted. Exploratory analyses are first used to examine the empirical mean, variance, and correlation structure of the data and to assess the nature of missingness. Subsequently, likelihood-based linear mixed-effects models are fitted for the continuous BPRS outcome, allowing for subject-specific trajectories and valid inference under a missing-at-random (MAR) assumption (Laird and Ware, 1982; Verbeke and Molenberghs, 2000).

For a binary cognitive outcome derived from the Clinical Dementia Rating Sum of Boxes (CDR-SB), both subject-specific and population-averaged modelling frameworks are considered. Random-effects logistic regression models are used to obtain subject-specific interpretations, while generalized estimating equations (GEE) are employed to estimate marginal effects. Because standard GEE is not generally valid under MAR dropout, inverse-probability weighted GEE is used to obtain MAR-consistent marginal inference (Robins et al., 1995; Fitzmaurice et al., 2011).

Finally, recognising that the MAR assumption cannot be verified empirically from the observed data, a sensitivity analysis is conducted for the continuous outcome using a pattern-mixture \(\delta\)-adjustment approach. This analysis evaluates the robustness of the substantive conclusions to plausible departures from MAR by allowing unobserved post-dropout outcomes to differ systematically from MAR-based predictions (Little, 1993; Molenberghs et al., 2008). Overall, these methods provide a principled framework for analysing longitudinal Alzheimer’s disease data in the presence of dropout and for assessing the sensitivity of conclusions to assumptions about the missing data mechanism.



# Methodology

## Missingness in the data

### Notation and missingness indicators

Let $i=1,\dots,N$ index patients and $j=0,\dots,J$ index scheduled annual visits ($j=0$ baseline). The continuous outcome of interest is the BPRS score $Y_{ij}$. Define the missingness indicator

$$
R_{ij} =
\begin{cases}
1, & \text{if } Y_{ij}\ \text{is observed},\
0, & \text{if } Y_{ij}\ \text{is missing.}
\end{cases}
$$

Let $R_i=(R_{i0},\dots,R_{iJ})$ denote the subject-specific observation pattern, and split the full outcome vector as $Y_i=(Y_i^{obs},Y_i^{mis})$.

### Description of missingness patterns

Missingness patterns were first characterised descriptively to determine whether dropout followed a monotone structure or whether intermittent missingness occurred. For each subject, the last observed visit was defined as $t_i = \max\{j : R_{ij}=1$

A subject was classified as exhibiting monotone dropout if $(R_{ij}=1$ for all $(j \le t_i$ and $R_{ij}=0$ for all $(j > t_i)$. Non-monotone (intermittent) missingness was defined as the presence of at least one missing observation followed by a later observed measurement i.e The pattern is non-monotone if there exist $j<k$ such that $R_{ij}=0 \ \text{and}\ R_{ik}=1$. The distribution of $(t_i)$, the proportion missing at each visit, and the frequency of intermittent gaps were summarised to assess the overall structure of missingness.

These classifications are summarised using (i) the distribution of $t_i$ (retention over time), (ii) the proportion missing at each visit $j$, and (iii) the frequency of intermittent gaps within $R_i$. In this Alzheimer cohort, missingness rises markedly with follow-up, so later-visit summaries are based on a progressively smaller subset of participants.

### Empirical assessment of dependence on observed data

To assess whether missingness was plausibly independent of observed outcomes (MCAR) or related to observed disease severity and covariates (supporting MAR), empirical summaries were examined. First, the relationship between baseline BPRS and the total number of missed follow-up visits was evaluated to assess whether initial psychiatric severity was associated with subsequent dropout. Second, visit-specific missingness rates were calculated and plotted to characterise the evolution of missingness over time and the degree of selection at later follow-up visits.

### Discrete-time model for dropout

To formalise the dependence of missingness on observed history, a discrete-time observation (dropout) model was fitted. Among subjects observed at visit $(j\)$, a next-visit missingness indicator was defined as:
$$
D_{i,j+1} = 1 - R_{i,j+1}.
$$

The conditional probability of dropout at the next visit was modelled using logistic regression:
$$
\Pr(D_{i,j+1}=1 \mid \mathcal{H}_{ij})
= \operatorname{logit}^{-1}\!\left(
\alpha_0 + \alpha_1 Y_{ij} + \alpha_2^\top X_i + \alpha_3 j
\right),
$$
where ${H}_{ij}$ denotes the observed outcome and covariate history up to visit $(j)$, and $(X_i)$ includes baseline demographic and clinical covariates. Evidence of dependence of $(D_{i,j+1})$ on observed outcomes or covariates was interpreted as evidence against MCAR and supportive of a MAR mechanism.

### Working assumptions and implications for analysis

Under the MAR assumption, the missingness mechanism satisfies:
$$
p(R_i \mid Y_i, X_i) = p(R_i \mid Y_i^{\mathrm{obs}}, X_i),
$$
so that, conditional on observed data, missingness does not depend on unobserved outcomes. Under this assumption and assuming distinct parameters for the outcome and missingness processes, likelihood-based inference using the observed-data likelihood is valid. Accordingly, MAR was adopted as the primary working assumption for the main longitudinal analyses.

Because MAR cannot be verified empirically and MNAR mechanisms where missingness depends on unobserved outcomes cannot be identified from the observed data alone, sensitivity analyses were conducted for the continuous outcome to assess the robustness of substantive conclusions to plausible departures from MAR.


## Longitudinal outcome models (continuous and binary)

### Notation and outcomes

Let $i=1,\ldots,N$ index subjects and $j=0,\ldots,J$ index scheduled annual visits, with $j=0$ denoting baseline. Let $t_{ij}$ denote follow-up time in years (here $t_{ij}=j$). The primary continuous outcome is the BPRS score $Y_{ij}$. A binary cognitive outcome is defined from the CDR-SB score by
$$
Z_{ij}=\mathbb{I}\!\left(\mathrm{CDRSB}_{ij}>10\right),
$$
where ${I}(\cdot)$ denotes the indicator function. Baseline covariates are collected in the vector $X_i$, and baseline disease-severity markers are denoted by $S_i$ (e.g., ${CDRSB}_{i0}$ and baseline biomarkers).

### Continuous outcome model: linear mixed-effects model (LMM)

To model the longitudinal evolution of BPRS while accounting for within-subject correlation and unbalanced follow-up, a linear mixed-effects model was specified as
$$
Y_{ij}
=
\beta_0
+
\beta_1 t_{ij}
+
\beta^\top X_i
+
\gamma^\top S_i
+
\eta\, t_{ij}\,\mathrm{CDRSB}_{i0}
+
b_{0i}
+
b_{1i} t_{ij}
+
\varepsilon_{ij}.
$$
Here, $\beta_0$ is the population intercept, $\beta_1$ is the average yearly change in BPRS, $\beta$ and $\gamma$ quantify associations with baseline covariates and baseline severity markers, and $\eta$ captures effect modification of time by baseline CDR-SB. The subject-specific random effects $(b_{0i},b_{1i})$ represent heterogeneity in baseline BPRS and progression rate and are assumed to follow a bivariate normal distribution with mean zero and covariance matrix $D$. Conditional on the random effects, the residuals $\varepsilon_{ij}$ are assumed independent and normally distributed with variance $\sigma^2$.

This model was chosen because it explicitly accounts for within-subject dependence, accommodates unbalanced follow-up due to dropout, and yields likelihood-based inference that is valid under a missing-at-random (MAR) mechanism when the mean and covariance structures are correctly specified.

### Binary outcome models

#### Subject-specific model: logistic generalized linear mixed model (GLMM)

For the binary outcome $Z_{ij}$, a subject-specific logistic mixed model was specified as
$$
\Pr\!\left(Z_{ij}=1 \mid b_{0i}\right)
=
\operatorname{logit}^{-1}\!\left(
\alpha_0
+
\alpha_1 t_{ij}
+
\alpha^\top X_i
+
\delta^\top S_i
+
b_{0i}
\right),
$$
where $b_{0i}\sim N(0,\tau^2)$ is a random intercept capturing within-subject correlation. This model yields subject-specific (conditional) odds ratios and, being likelihood-based, provides valid inference under MAR given standard ignorability conditions. A random-intercept specification was adopted as a parsimonious yet adequate representation of within-subject dependence for the binary outcome.


#### Marginal model: generalized estimating equations (GEE)

To estimate population-averaged effects for the binary outcome, a marginal logistic mean model was considered:
$$
\Pr\!\left(Z_{ij}=1\right)
=
\operatorname{logit}^{-1}\!\left(
\theta_0
+
\theta_1 t_{ij}
+
\theta^\top X_i
+
\kappa^\top S_i
\right),
$$
with within-subject correlation handled through a working correlation structure. Standard GEE yields consistent regression estimates under MCAR, but it is not generally valid under MAR dropout when the probability of observation depends on observed history. Given the strong time-dependent dropout and its association with observed severity and covariates in these data, a MAR-valid marginal approach was required.


### MAR-valid marginal inference: inverse-probability weighted GEE (IPW-GEE)

To obtain marginal estimates valid under MAR, inverse-probability weighted GEE was employed. Let $R_{ij}$ denote the indicator that $Z_{ij}$ is observed, and let $\mathcal{H}_{i,j-1}$ denote the observed history up to visit $j-1$. An observation model was fitted to estimate the probability of being observed at visit $j$:
$$
\widehat{\pi}_{ij}
=
\Pr\!\left(R_{ij}=1 \mid \mathcal{H}_{i,j-1}\right).
$$
Inverse-probability weights were then constructed as
$$
w_{ij}=\frac{1}{\widehat{\pi}_{ij}},
$$
and incorporated into the GEE estimating equations for the marginal model. Under correct specification of the observation model, this weighting adjusts for the selection induced by MAR dropout and yields consistent population-averaged regression estimates.

IPW-GEE was selected as the marginal approach because missingness depended on observed history and follow-up time (inconsistent with MCAR), the scientific estimand was population-averaged, and inverse-probability weighting provides a transparent and principled correction for MAR dropout while retaining the marginal interpretation of GEE.

### Model selection and primary analysis

For the continuous outcome (BPRS), the linear mixed-effects model was selected as the primary analysis model. This choice is motivated by the continuous nature of the outcome, the presence of strong within-subject correlation, and the unbalanced follow-up arising from monotone dropout. The likelihood-based mixed model provides subject-specific trajectories, accommodates heterogeneity in baseline level and progression rate through random effects, and yields valid inference under the MAR assumption supported by the missingness analyses.

For the binary outcome derived from CDR-SB, both subject-specific and marginal modelling frameworks were considered to address different scientific targets. The logistic generalized linear mixed model (GLMM) was used to obtain subject-specific effects, which are appropriate when interest lies in individual-level disease progression. In contrast, the marginal model was used to estimate population-averaged effects. Because standard GEE is not generally valid under MAR dropout and missingness in this cohort was shown to depend on observed history and follow-up time, inverse-probability weighted GEE (IPW-GEE) was selected as the primary marginal approach. IPW-GEE yields consistent population-averaged estimates under MAR by correcting for the selection induced by dropout.

Together, these modelling choices ensure that inference for both continuous and binary outcomes is aligned with the observed data structure, the missingness mechanism, and the scientific objectives of the study.


### MAR-valid model choice and justification (continuous outcome)

For the continuous outcome, the Brief Psychiatric Rating Scale (BPRS), a likelihood-based linear mixed-effects model (LMM) was selected as the final analysis model. The LMM is appropriate because BPRS is measured on a continuous scale and repeatedly over time, inducing within-subject correlation that is naturally accommodated through random effects. In addition, the model allows for subject-specific trajectories by incorporating random intercepts and slopes, thereby capturing heterogeneity in both baseline psychiatric symptom severity and rates of change over follow-up.

Under the missing-at-random (MAR) assumption supported by the missingness diagnostics, likelihood-based estimation in the LMM yields valid inference using the observed-data likelihood and appropriately handles the unbalanced follow-up arising from monotone dropout. Alternative approaches were considered, but the LMM was selected as the primary and final model because it aligns with the scientific objective of modelling continuous BPRS trajectories, accommodates within-subject dependence and heterogeneity, and provides principled, MAR-valid inference without the need for additional weighting or imputation.

### MAR-valid model choice and justification (binary outcome)

For the binary outcome derived from the continuous CDR-SB score, defined as $Z_{ij}=\mathbb{I}(\mathrm{CDRSB}_{ij}>10)$, a logistic generalized linear mixed model (GLMM) was selected as the primary analysis model. This choice is motivated by the binary nature of $Z_{ij}$ and the longitudinal design, which induces within-subject correlation across repeated measurements. The GLMM accounts for this dependence through subject-specific random effects and allows individuals to have different baseline propensities for $Z_{ij}=1$ (and, if included, different progression rates over time).

The model is specified as
$$
\Pr(Z_{ij}=1\mid b_{0i})
=
\operatorname{logit}^{-1}\!\left(
\alpha_0+\alpha_1 t_{ij}+\alpha^\top X_i+\delta^\top S_i+b_{0i}
\right),
$$
with $b_{0i}\sim N(0,\tau^2)$. As a likelihood-based method, the GLMM yields valid inference under a missing-at-random (MAR) mechanism, provided that the outcome model is correctly specified and the missingness process is ignorable (i.e., depends only on observed data and has distinct parameters from the outcome model). Given the observed monotone dropout and evidence that missingness depends on observed covariates and history, the MAR assumption is a reasonable primary working assumption, making the GLMM an appropriate MAR-valid modelling choice for the binary endpoint.


# Results

### Missingness structure and missingness mechanism (BPRS)

### Continuous outcome model (BPRS): linear mixed-effects model under MAR

### Binary outcome model: logistic GLMM under MAR

### Agreement between exploratory patterns (Assignment 1) and fitted models

### Sensitivity analysis for MNAR (continuous outcome)


###  Proportion missing by time

```{r}

library(data.table)
library(ggplot2)

# Ensure data.table
setDT(alzheimer_long)
setorder(alzheimer_long, patid, time)

alzheimer_long[, time := time - 1L]


# --- Choose outcome to study (start with BPRS)
outcome <- "bprs"

# 1) Create missingness indicator R_ij for chosen outcome
alzheimer_long[, R := as.integer(!is.na(get(outcome)))]

# 2) Proportion missing by time
miss_by_time <- alzheimer_long[, .(
  n_total = .N,
  n_obs   = sum(R == 1),
  n_miss  = sum(R == 0),
  prop_miss = mean(R == 0)
), by = time][order(time)]

miss_by_time
```


```{r}
# --- Choose outcome to study (start with BPRS)
outcome <- "cdrsb"

# 1) Create missingness indicator R_ij for chosen outcome
alzheimer_long[, R := as.integer(!is.na(get(outcome)))]

# 2) Proportion missing by time
miss_by_time1 <- alzheimer_long[, .(
  n_total = .N,
  n_obs   = sum(R == 1),
  n_miss  = sum(R == 0),
  prop_miss = mean(R == 0)
), by = time][order(time)]

miss_by_time1
```


```{r}
ggplot(miss_by_time, aes(x = time, y = prop_miss)) +
  geom_line(color = "blue", size = 1) + geom_point(color = "blue", size = 2) +
  scale_x_continuous(breaks = 0:6) +
  labs(title = paste0("Proportion missing in ", outcome, " by visit"),
       x = "Time (0=baseline, ..., 6)", y = "Proportion missing") +
  theme_minimal()
```


### Subject-level pattern features: t_i, monotone vs non-monotone, gaps

```{r}
# 3) Subject-level pattern features: t_i, monotone vs non-monotone, gaps
# Helper: monotone dropout if once missing => always missing
is_monotone <- function(Rvec) all(cummin(Rvec) == Rvec)

pattern_dt <- alzheimer_long[, {
  Rvec <- R
  # last observed time (t_i); if all missing, set NA (rare if baseline always observed)
  if (all(Rvec == 0)) {
    list(t_i = NA_integer_, complete = FALSE, monotone = TRUE, non_monotone = FALSE, n_gaps = NA_integer_) 
  } else {
    t_i <- max(time[Rvec == 1])
    n_gaps <- sum(Rvec == 0 & time < t_i)   # missing before last observed => intermittent gap(s)
    mono <- is_monotone(Rvec)
    list(
      t_i = t_i,
      complete = all(Rvec == 1),
      monotone = mono,
      non_monotone = !mono,
      n_gaps = n_gaps
    )
  }
}, by = patid]

pattern_dt[, pattern := fifelse(complete, "Complete",
                        fifelse(monotone, "Monotone dropout", "Non-monotone (intermittent)"))]

# Pattern counts
pattern_dt[, .N, by = pattern]

```


###Distribution of t_i (retention / dropout timing)

```{r}
# 4) Distribution of t_i (retention / dropout timing)
ti_dist <- pattern_dt[!is.na(t_i), .N, by = t_i][order(t_i)]
ti_dist

ggplot(pattern_dt[!is.na(t_i)], aes(x = factor(t_i))) +
  geom_bar(fill = "blue") +
  labs(title = paste0("Dropout timing: distribution of t_i (", outcome, ")"),
       x = "t_i (last observed visit)", y = "Number of patients") +
  theme_minimal()
```



### Intermittent missingness summary
```{r}
# 5) Intermittent missingness summary
pattern_dt[, .(
  n_patients = .N,
  n_non_monotone = sum(pattern == "Non-monotone (intermittent)"),
  prop_non_monotone = mean(pattern == "Non-monotone (intermittent)"),
  mean_gaps_non_mon = mean(n_gaps[pattern == "Non-monotone (intermittent)"], na.rm = TRUE)
)]
```


```{r}
# 6) Show example R_i sequences for a few non-monotone patients
example_ids <- pattern_dt[pattern == "Non-monotone (intermittent)"][1:5, patid]

Ri_examples <- alzheimer_long[patid %in% example_ids,
                              dcast(.SD, patid ~ time, value.var = "R")]

Ri_examples
```

### Missingness Conclusion

Missingness in BPRS increases markedly with follow-up, reaching its highest level at the final visit, indicating substantial attrition over time. Examination of subject-specific observation patterns shows that missingness is entirely monotone: participants who miss a BPRS measurement do not re-enter the study at later visits. Specifically, 742 participants exhibit monotone dropout and 511 are complete cases; no intermittent (non-monotone) missingness is observed. The distribution of the last observed visit $t_i$ confirms dropout occurs throughout follow-up, so later-visit summaries represent a progressively selected subset of patients.

Because follow-up is incomplete, the data are unbalanced, with progressively fewer observations at later visits. This makes methods that require complete data (such as repeated-measures ANOVA) inappropriate and potentially biased. The missingness pattern is monotone, consistent with dropout, so it is reasonable to work under a MAR framework (at least as a starting assumption) and use models that properly accommodate dropout and within-subject dependence. In this setting, likelihood-based mixed-effects models are recommended such as linear mixed models for the continuous outcome (BPRS) and GLMMs for the binary outcome because they can use all available data despite unequal numbers of measurements per patient, explicitly account for within-patient correlation, and provide valid inference under MAR when the mean and covariance structures are correctly specified.


### Empirical signals of informative missingness

Because BPRS is measured repeatedly, a second goal is to evaluate whether missingness is associated with observed disease severity/history (evidence against MCAR and supportive of MAR as a working assumption). Two descriptive comparisons are used:

* **Baseline severity vs missingness burden.** Let $M_i=\sum_{j=1}^J (1-R_{ij})$ be the number of missed follow-up BPRS measurements. The relationship between baseline BPRS ($Y_{i0}$) and $M_i$ is examined to assess whether those with worse baseline psychiatric symptoms are more likely to miss subsequent visits.

Figure X shows the relationship between baseline BPRS ($Y_i0$) and the number of missed follow-up visits ($M_i$). There is a clear positive association: patients with higher baseline BPRS scores tend to miss more subsequent follow-up measurements. In particular, individuals with low baseline BPRS are predominantly complete or have few missed visits, whereas those with higher baseline psychiatric symptom severity exhibit substantially greater missingness burden. This pattern indicates that dropout is related to observed disease severity, providing strong evidence against a MCAR mechanism and supporting MAR as a reasonable working assumption for the missingness process. Because missingness depends on observed baseline severity, likelihood-based mixed-effects models that are valid under MAR are appropriate for primary longitudinal analyses.

```{r baseline-severity-missingness-burden}
library(data.table)
library(ggplot2)

setDT(alzheimer_long)
setorder(alzheimer_long, patid, time)

# Ensure time is 0..6 (if yours is 1..7, uncomment next line)
# alzheimer_long[, time := time - 1L]

# Missingness indicator for BPRS
alzheimer_long[, R_bprs := as.integer(!is.na(bprs))]

# Subject-level: baseline BPRS and number of missed follow-ups (exclude baseline j=0)
burden_dt <- alzheimer_long[, .(
  bprs0 = bprs[time == 0][1],
  M_i   = sum(1 - R_bprs[time >= 1])  # missed follow-ups only
), by = patid]

# Quick summary table
burden_dt[, .(
  n = .N,
  mean_M = mean(M_i, na.rm = TRUE),
  sd_M   = sd(M_i, na.rm = TRUE),
  min_M  = min(M_i, na.rm = TRUE),
  max_M  = max(M_i, na.rm = TRUE)
)]

# Scatter + smooth trend: baseline severity vs missingness burden
ggplot(burden_dt, aes(x = bprs0, y = M_i)) +
  geom_jitter(color = "blue", width = 0.2, height = 0.1, alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Baseline BPRS vs missing follow-up burden (M_i)",
       x = "Baseline BPRS (Y_i0)", y = "Missed follow-up visits (M_i)") +
  theme_minimal()

```


* **Visit-specific missingness rates.** The proportion missing at each visit $j$,

  $$
  \widehat{p}*j=\frac{1}{N}\sum*{i=1}^N (1-R_{ij}),
  $$

  is plotted over time to characterise the growth of missingness with follow-up (and, indirectly, the degree of selection in later years). 
  
Figure Y displays the visit-specific missingness rates $𝑝^𝑗$ for BPRS over time. Missingness increases monotonically with follow-up, from essentially zero at baseline to approximately 60% by the final visit. This steady rise indicates substantial attrition and increasing selection among participants who remain under observation at later time points. Consequently, summaries at later visits are based on progressively smaller and more selected subsets of patients, reinforcing the need for modeling approaches that can accommodate unbalanced longitudinal data and informative dropout.

The strong time dependence of $𝑝^𝑗$ further rules out MCAR and motivates likelihood-based mixed-effects models that remain valid under MAR.

These summaries establish whether the data resemble monotone dropout, whether intermittent missingness occurs, and whether observed severity/history is linked to being missing—key inputs for choosing MAR-valid modelling strategies later.



```{r visit-specific-missingness-rates}
# Visit-specific missingness for BPRS
miss_by_visit <- alzheimer_long[, .(
  N = uniqueN(patid),
  n_obs = sum(R_bprs == 1),
  n_miss = sum(R_bprs == 0),
  p_hat = mean(R_bprs == 0)
), by = time][order(time)]

miss_by_visit

ggplot(miss_by_visit, aes(x = time, y = p_hat)) +
  geom_line(color = "blue") + geom_point(color = "blue") +
  scale_x_continuous(breaks = 0:6) +
  labs(title = "Visit-specific missingness rate for BPRS",
       x = "Visit (time)", y = expression(hat(p)[j])) +
  theme_minimal()
```


### Working assessment of MCAR vs MAR, and why MNAR cannot be “diagnosed” from observed data

To formalise the dependence of missingness on observed history, a discrete-time observation model is fitted. Define a “next-visit missingness” indicator for those observed at visit $j$:

$$
D_{i,j+1} = 1-R_{i,j+1}.
$$

A logistic regression (discrete-time hazard) model is used:

$$
\Pr(D_{i,j+1}=1 \mid \mathcal{H}*{ij})
= \operatorname{logit}^{-1}!\Big(
\alpha_0 + \alpha_1 Y*{ij} + \alpha_2^\top X_i + \alpha_3 j
\Big),
$$


where $\mathcal{H}*{ij}$ is the observed history up to $j$ and $X_i$ contains baseline covariates (e.g., sex, age, BMI, ADL, job status, living situation, baseline CDR-SB, and other baseline markers as available). A clear dependence of $D*{i,j+1}$ on observed outcomes/covariates argues against MCAR and supports MAR as a reasonable primary assumption.

The MAR mechanism is stated as

$$
p(R_i \mid Y_i, X_i)=p(R_i \mid Y_i^{obs}, X_i),
$$

whereas MNAR allows dependence on $Y_i^{mis}$. Since MNAR is defined through unobserved outcomes, it cannot be identified from the observed data alone; therefore, departures from MAR are addressed through sensitivity analysis rather than “testing” MNAR.



```{r discrete-time-dropout-model}
library(data.table)

setDT(alzheimer_long)
setorder(alzheimer_long, patid, time)

# Ensure time is 0..6 (if yours is 1..7, uncomment next line)
# alzheimer_long[, time := time - 1L]

# 1) Missingness indicator for current visit (R_ij)
alzheimer_long[, R := as.integer(!is.na(bprs))]

# 2) Construct next-visit missingness indicator:
#    D_{i,j+1} = 1 - R_{i,j+1}
alzheimer_long[, D_next := 1L - shift(R, type = "lead"), by = patid]

# 3) Keep only rows where:
#    - current visit is observed (R_ij = 1), so the "risk set" at time j
#    - next visit exists (time <= 5 if visits are 0..6)
obs_model_dt <- alzheimer_long[
  R == 1 & time <= 5,
  .(patid, time, D_next,
    y_ij = bprs,                 # Y_ij = current observed BPRS
    age, sex, bmi, job, adl, wzc,
    cdrsb0, abpet0, taupet0, edu)
]

# Optional: drop rows where D_next is NA (should only happen at last visit)
obs_model_dt <- obs_model_dt[!is.na(D_next)]

# 4) Fit discrete-time hazard (logistic regression)
dropout_glm <- glm(
  D_next ~ y_ij + age + sex + bmi + job + adl + wzc + cdrsb0 + abpet0 + taupet0 + edu + time,
  data = obs_model_dt,
  family = binomial
)

summary(dropout_glm)

# Optional: odds ratios (more interpretable)
OR <- exp(cbind(OR = coef(dropout_glm), confint(dropout_glm)))
OR

```

The discrete-time observation (dropout) model provides strong evidence that missingness depends on observed history and observed covariates, thereby ruling out MCAR as a plausible mechanism. The probability of being missing at the next visit increases sharply with follow-up time (odds ratio \(=3.27\) per year, \(p<2\times10^{-13}\)), indicating substantial time-dependent attrition. Several baseline clinical characteristics are also strongly associated with next-visit missingness: ADL is highly predictive (odds ratio \(=0.15\), \(p<2\times10^{-16}\)), nursing home residence markedly increases the risk of dropout (odds ratio \(=6.05\), \(p<2\times10^{-16}\)), and baseline CDR-SB shows a significant positive association with dropout (odds ratio \(=1.05\) per unit, \(p=4.3\times10^{-7}\)). 

In contrast, the current observed BPRS level \(Y_{ij}\) is not statistically significant (\(p=0.15\)), although its estimated effect is in the positive direction. Overall, because missingness clearly depends on observed covariates and visit time, the data are inconsistent with MCAR and support MAR as a reasonable primary working assumption.

Although MAR is supported by the observed data, an MNAR mechanism—where missingness depends on unobserved future outcomes cannot be ruled out. By definition, MNAR involves dependence on unobserved responses and therefore cannot be diagnosed using the observed data alone; potential departures from MAR must instead be investigated through sensitivity analyses.


### MAR assumption used for primary analyses (and link to later models)

For primary modelling, missingness is treated as ignorable under MAR (with distinct parameters for the outcome and missingness processes), so inference can proceed using the observed-data likelihood for the continuous outcome models. This aligns with the modelling assumptions used for BPRS longitudinal mixed modelling in this dataset. 

In addition, the fitted observation model above yields estimated observation probabilities
$$
\widehat{\pi}*{ij}=\Pr(R*{ij}=1 \mid \mathcal{H}*{i,j-1}),
$$

which can be converted into inverse-probability weights $w*{ij}=1/\widehat{\pi}_{ij}$ for MAR-valid weighted marginal analyses when needed.



# Question2. In the first assignment, linear mixed models have been considered. Select one of them

```{r}
library(lme4 )
fm_adj <- bprs ~ time + sex + 
  age + bmi + job + adl + wzc +  cdrsb0 + 
   time:cdrsb0 +
   1 + (time | patid) + (1 | trial)
lmer_reduced_mean_structure <- lmer(fm_adj,
  data = alzheimer_long,
  na.action = na.exclude
)

summary(lmer_reduced_mean_structure)
```



### Question 3 For the binary outcome, consider a random-effects as well as a marginal model.

#### Marginal Model

```{r}

# id_vars <- c("patid", "trial", "sex", "age", 
# "edu", "bmi", "inkomen", "job", "adl", "wzc")
          



alzheimer_long <- melt(alzheimer_data,
  id.vars = id_vars,
  measure.vars = patterns(
    cdrsb = "^cdrsb[0-6]",
    bprs = "^bprs[0-6]",
    abpet = "^abpet[0-6]",
    taupet = "^taupet[0-6]"
  ),
  variable.name = "time",
  value.name = c("cdrsb", "bprs", "abpet", "taupet"),
  variable.factor = FALSE
)

alzheimer_long[, time := as.integer(time)]

alzheimer_long[, CDRSBbin := ifelse(cdrsb > 10, 1, 0)]
#alzheimer_long[, time := time - 1]


# Display the first few rows of the new long-format data
#kable(head(alzheimer_long))
```



```{r}
library(geepack)
# Ensure data is sorted for GEE
setorder(alzheimer_long, patid, time)

gee_model <- geeglm(
    CDRSBbin ~ time + sex + age + bmi +
     adl + abpet + taupet + 
     time:abpet + time:taupet,
    id = patid,
    data = alzheimer_long,
    family = binomial(link = "logit"),
    corstr = "ar1",
    std.err = "san.se" # Robust standard errors
)

summary(gee_model)
```



#### Random Effects model

```{r}
alzheimer_long <- alzheimer_long |>
  dplyr::mutate(
    time_c   = time   - mean(time,   na.rm = TRUE),
    age_c    = age    - mean(age,    na.rm = TRUE),
    bmi_c    = bmi    - mean(bmi,    na.rm = TRUE),
    adl_c    = adl    - mean(adl,    na.rm = TRUE),
    abpet_c  = abpet  - mean(abpet,  na.rm = TRUE),
    taupet_c = taupet - mean(taupet, na.rm = TRUE)
  )

ctrl <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
glmm_center <- glmer(
  CDRSBbin ~ time_c + sex + age_c + bmi_c + adl_c +
    abpet_c + taupet_c + time_c:taupet_c +
    (time_c | patid),
  data   = alzheimer_long,
  family = binomial(link = "logit"),
  nAGQ   = 1,
  control = ctrl
)
summary(glmm_center)
```


### Question 3  For each of these three models, present one version that is valid under MAR. While thereare several possible solutions in each of the three cases, it is explicitly the idea to select one of them and then argue why you have made this choice


1. Continuous outcome (BPRS): MAR-valid Linear Mixed Model (likelihood-based)
MAR is valide becauseLikelihood-based LMM uses the observed-data likelihood and is valid under MAR (with correct mean/covariance).

```{r mar-lmm-bprs}
library(data.table)
library(lme4)
library(lmerTest)

setDT(alzheimer_long)
setorder(alzheimer_long, patid, time)

# Ensure time is 0..6 (uncomment if yours is 1..7)
# alzheimer_long[, time := time - 1L]

# (adjust covariates to match your final selection)
lmm_mar <- lmer(
  bprs ~ time + age + sex + bmi + job + adl + wzc + cdrsb0 + time:cdrsb0 + (time | patid),
  data = alzheimer_long,
  REML = FALSE
)

summary(lmm_mar)
```

2. Binary outcome (CDRSBbin): MAR-valid Random-effects logistic model (GLMM).MAR-valid Because Likelihood-based GLMM is valid under MAR (again, under correct model + ignorability)

```{r mar-glmm-cdrsbbin}
library(lme4)

# Make sure CDRSBbin exists
alzheimer_long[, CDRSBbin := ifelse(cdrsb > 10, 1L, 0L)]

# Random intercept logistic model (you can use (time|patid) if needed)
glmm_mar <- glmer(
  CDRSBbin ~ time_c + sex + age_c + bmi_c + adl_c +
    abpet_c + taupet_c + time_c:taupet_c +
    (time_c | patid),
  data = alzheimer_long,
  family = binomial,
  nAGQ = 1
)

summary(glmm_mar)
```

3. Weighted GEE
### Problem: why standard GEE is not generally MAR-valid

Standard GEE yields consistent estimation under **MCAR** (and in some special cases), but it is **not generally valid under MAR** when dropout depends on observed history. In our data, missingness increases with follow-up and is associated with observed severity and covariates, so a MAR-valid marginal approach is required.

### MAR-valid version: Weighted GEE (IPW-GEE)

To obtain a marginal model that is valid under MAR, we use **inverse-probability weighted GEE (IPW-GEE)**. First, an observation (missingness) model is fitted to estimate the probability that subject \(i\) is observed at visit \(j\), given the observed history up to the previous visit:

$$
\widehat{\pi}_{ij} = \Pr\!\left(R_{ij}=1 \mid \mathcal{H}_{i,j-1}\right).
$$

Inverse probability weights are then constructed as:

$$
w_{ij}=\frac{1}{\widehat{\pi}_{ij}}.
$$

Next, the marginal mean model for the binary outcome is fitted:

$$
\Pr(Y_{ij}=1)
= \operatorname{logit}^{-1}\!\left(\beta_0+\beta_1\,\text{time}_{ij}+\beta^\top X_i\right),
$$

using GEE with weights \(w_{ij}\).

### Why IPW-GEE is MAR-valid

If the observation model for \(\pi_{ij}\) is correctly specified, the inverse-probability weights correct for the selection induced by MAR dropout. Therefore, the weighted estimating equations yield **consistent marginal (population-averaged) estimates under MAR**.

**Justification (choice of method):**

I use IPW-weighted GEE because standard GEE is not generally valid under MAR dropout; inverse probability weights based on the fitted observation model yield consistent marginal estimates under MAR.


```{r}
# ============================================================
# MAR-valid Weighted GEE (IPW-GEE) for CDRSBbin
#   - Step 1: fit a dropout/observation model (discrete-time)
#   - Step 2: build inverse-probability-of-observation weights
#   - Step 3: fit weighted GEE (geeglm)
# ============================================================

library(data.table)
library(geepack)

setDT(alzheimer_long)
setorder(alzheimer_long, patid, time)

# If your time is 1..7, convert to 0..6
# alzheimer_long[, time := time - 1L]

# -----------------------------
# Outcome + missingness indicator
# -----------------------------
alzheimer_long[, CDRSBbin := ifelse(!is.na(cdrsb) & cdrsb > 10, 1L,
                            ifelse(!is.na(cdrsb), 0L, NA_integer_))]

# Observed indicator for continuous cdrsb (same missingness as CDRSBbin)
alzheimer_long[, R := as.integer(!is.na(cdrsb))]

# -----------------------------
# Step 1: Discrete-time dropout model
# D_{i,j+1} = 1 - R_{i,j+1}, among those observed at time j
# -----------------------------
alzheimer_long[, D_next := 1L - shift(R, type = "lead"), by = patid]

# Risk set: observed at time j, and j+1 exists (if time 0..6, last j is 5)
obs_dt <- alzheimer_long[R == 1 & time <= 5,
                         .(patid, time, D_next,
                           y_ij = cdrsb,   # observed history at time j
                           age, sex, bmi, job, adl, wzc, cdrsb0, edu)]

obs_dt <- obs_dt[!is.na(D_next)]

# Fit dropout model (you can drop covariates that cause separation if needed, e.g., job)
dropout_glm <- glm(
  D_next ~ time + y_ij + age + sex + bmi + job + adl + wzc + cdrsb0 + edu,
  data = obs_dt,
  family = binomial
)

# Predict conditional probability of being observed at next visit:
# p_obs_next(i,j) = P(R_{i,j+1}=1 | history up to j, R_{ij}=1)
obs_dt[, p_drop_next := predict(dropout_glm, type = "response")]
obs_dt[, p_obs_next  := pmax(1 - p_drop_next, 1e-6)]  # guard against 0

# Create table mapping p_obs_next to the NEXT visit (time+1)
p_next <- obs_dt[, .(patid, time = time + 1L, p_obs_step = p_obs_next)]

# -----------------------------
# Step 2: Build cumulative observation probabilities pi_it
# pi_it = Π_{k=1..t} p_obs_step(i,k)
# where p_obs_step(i,k) = P(observed at time k | observed at time k-1, history)
# -----------------------------
w_dt <- unique(alzheimer_long[, .(patid, time)])
setorder(w_dt, patid, time)

# Merge step-probabilities onto each (patid, time)
w_dt <- merge(w_dt, p_next, by = c("patid", "time"), all.x = TRUE)

# Baseline time=0: set step prob = 1 (always observed at baseline in your study)
w_dt[time == 0, p_obs_step := 1]

# For visits with no model-based probability (should be only impossible cases), set NA to 1
w_dt[is.na(p_obs_step), p_obs_step := 1]

# Cumulative probability of being observed up to time t
w_dt[, pi_obs := cumprod(p_obs_step), by = patid]

# Stabilized weights (recommended): numerator = marginal P(observed at time)
# Compute marginal observation rate by time:
marg_obs <- alzheimer_long[, .(p_marg_obs = mean(R == 1, na.rm = TRUE)), by = time]
w_dt <- merge(w_dt, marg_obs, by = "time", all.x = TRUE)
w_dt[is.na(p_marg_obs), p_marg_obs := 1]

# Stabilized cumulative numerator
w_dt[, num_pi := cumprod(p_marg_obs), by = patid]

# Stabilized IPW
w_dt[, w_ipw := num_pi / pi_obs]

# Optional: truncate extreme weights for stability (common)
cap <- quantile(w_dt$w_ipw, 0.99, na.rm = TRUE)
w_dt[w_ipw > cap, w_ipw := cap]

# Attach weights back to main long data
alzheimer_long <- merge(alzheimer_long, w_dt[, .(patid, time, w_ipw)], by = c("patid", "time"), all.x = TRUE)

# -----------------------------
# Step 3: Weighted GEE (marginal model) for observed CDRSBbin
# -----------------------------
gee_dt <- alzheimer_long[!is.na(CDRSBbin)]

wgee_mar <- geeglm(
  CDRSBbin ~ time + age + sex + bmi + job + adl + wzc + cdrsb0 + edu,
  id = patid,
  data = gee_dt,
  family = binomial,
  corstr = "exchangeable",
  weights = w_ipw
)

summary(wgee_mar)

```


### Question 5 Examine whether the results obtained from data exploration in the first assignment arein agreement with the results obtained from some of these models. As an example: isthe empirical variance function of Assignment 1 sufficiently similar to the fitted variancefunction in the version of the linear mixed model considered here? Would it be a problem if they differ, and why (not)?

```{r}
#| label: fig-missingness-analysis
#| fig-cap: "Missingness Analysis: Baseline BPRS by Number of Missed Visits (left) and Percentage Missing Over Time (right)"
#| fig.width: 8.5
#| fig.height: 3.5

bprs_cols <- paste0("bprs", 0:6)
alzheimer_data[, n_missing := rowSums(is.na(.SD)), .SDcols = bprs_cols]

# Group by the number of missing visits
# and calculate summary stats for baseline BPRS
dropout_summary <- alzheimer_data[!is.na(bprs0), .(
  mean_bprs0 = mean(bprs0),
  median_bprs0 = median(bprs0),
  visit_n_patients = .N
), by = n_missing][order(n_missing)]



missing_by_time <- alzheimer_long[, .(
  n_patients = uniqueN(patid),
  n_obs = sum(!is.na(bprs))
), by = time][order(time)]

missing_by_time[, pct_missing := round(100 *
                                         (n_patients - n_obs)/
                                         n_patients, 1)]



boxplot_meanbprso0 <- ggplot(alzheimer_data[!is.na(bprs0)], 
                             aes(x = factor(n_missing),
                                 y = bprs0)) +
  geom_boxplot(aes(fill = factor(n_missing)),
               alpha = 0.7) +
  labs(
    title = "Baseline BPRS Score by
    Number of Missed Visits",
    x = "Number of Missed 
    BPRS Measurements",
    y = "Baseline BPRS 
    Score (bprs0)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

missing_by_time_plot <- ggplot(missing_by_time, 
                               aes(x = time, y = pct_missing)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue", size = 2) +
  labs(
    title = "Percentage of Missing BPRS \n Measurements Over Time",
    x = "Time (Years)",
    y = "Percentage of Missing BPRS Measurements (%)"
  ) +
  theme_minimal()

grid.arrange(boxplot_meanbprso0,
             missing_by_time_plot,
             ncol = 2
             )
```

```{r}
#| label: loess-smoothing
#  LOESS model

loess_fit <- loess(bprs ~ time,
  data = alzheimer_long[!is.na(bprs)],
  na.action = na.omit
)

# Score the data to get predictions (equivalent to `score data=...`)

bprs_smoothed <- predict(loess_fit, newdata = alzheimer_long)

# Create and sort the output data

loess_results <- alzheimer_long[!is.na(bprs), .(patid, time, bprs, sex)]
loess_results[, predicted_bprs := predict(loess_fit, newdata = .SD)]

# Sort the results by time
setorder(loess_results, time)


# Visualize 
simple_avg_plot <- ggplot(alzheimer_long, aes(x = time, y = bprs)) +
  geom_point(alpha = 0.1, color = "gray") + # Plot raw data points transparently
  geom_line(
    data = loess_results,
    aes(y = predicted_bprs), color = "blue", size = 1
  ) +
  labs(
    title = "LOESS Smoothed Trend of BPRS Over Time",
    x = "Time (Years)",
    y = "BPRS Score"
  ) +
  theme_minimal()

simple_avg_plot
```


```{r}
#| label: fig-loess-by-categorical-vars
#| fig-width: 10.3
#| fig-height: 8
#| fig-cap: "LOESS Smoothed BPRS Over Time by Subgroups. Overall average (top),  sex, education, job status, living situation, and clinical center."


library(gridExtra)

subgroup_vars <- c("sex", 
                   "edu", 
                   "job",
                   "wzc", 
                   "trial")

# factor trial

alzheimer_long[, trial := factor(as.character(trial))]

mycolors <- c(
  "#E41A1C", "#377EB8", "#4DAF4A",
  "#984EA3", "#FF7F00", "#FFFF33",
  "#A65628", "#1B9E77", "#D95F02",
  "#7570B3", "#E7298A", "#66A61E",
  "#E6AB02", "#A6761D", "#7FC97F",
  "#BEAED4", "#FDC086", "#FFFF99",
  "#386CB0", "#F0027F", "#BF5B17",
  "#FBB4AE", "#B3CDE3", "#CCEBC5",
  "#DECBE4", "#FED9A6", "#FFFFCC", 
  "#E5D8BD"
)
# Titles for each plot
plot_titles <- c(
  "sex" = "LOESS Smoothed BPRS by Sex",
  "edu" = "LOESS Smoothed BPRS by Education",
  "job" = "LOESS Smoothed BPRS by Job Status",
  "wzc" = "LOESS Smoothed BPRS by Living Situation",
  "trial" = "LOESS Smoothed BPRS by Clinical Center",
  "taupet0" = "LOESS Smoothed BPRS by Baseline Tau PET"
)

#alzheimer_long[, taupet0 := factor(as.character(taupet0))]
# List to store plots
plot_list <- list()

# Loop to create plots
for (var in subgroup_vars) {
  p <- ggplot(
    alzheimer_long[!is.na(bprs)],
    aes_string(x = "time", y = "bprs", color = var)
  ) +
    geom_smooth(method = "loess", se = FALSE) +
    labs(
      title = plot_titles[var],
      x = "Time (Years)",
      y = "BPRS Score"
    ) +
    theme_minimal() +
    scale_color_manual(
      name = "",
      values =mycolors
    ) +
    theme(legend.position = "bottom")


  if (var == "trial") {
    p <- p + guides(color = guide_legend(ncol = 10, byrow = TRUE))
  }
  plot_list[[var]] <- p
}

names(plot_list) <- subgroup_vars

trial_centers_plot <- plot_list[["trial"]]
plot_list[["trial"]] <- NULL
# Arrange the plots in a grid
plot_list[["simple_avg_plot"]] <- simple_avg_plot


# create ordered grob list with overall plot first
glist <- plot_list[c(
  "simple_avg_plot",
  setdiff(
    names(plot_list),
    "simple_avg_plot"
  )
)]
glist <- unname(glist)


ncol <- 2
n_other <- length(glist) - 1
nrows <- 1 + ceiling(n_other / ncol)
layout_mat <- matrix(NA_integer_, nrow = nrows, ncol = ncol)
layout_mat[1, ] <- 1L
idx <- 2L
for (r in 2:nrows) {
  for (c in 1:ncol) {
    if (idx <= length(glist)) {
      layout_mat[r, c] <- idx
      idx <- idx + 1L
    }
  }
}

## add figure 2 caption
bottom_caption <- grid::textGrob(
  "Figure 2: LOESS Smoothed BPRS Over Time by Subgroups.
  Overall average (top),  sex, education, job status, 
  living situation, and clinical center.",
  gp = grid::gpar(fontsize = 10),
  x = 0.5,
  hjust = 0.5
)

grid.arrange(
  grobs = glist,
  layout_matrix = layout_mat
  #bottom = bottom_caption
)
```

The exploratory analysis in Assignment 1 indicated (i) a clear increase in the mean BPRS trajectory over time, (ii) strong within-subject dependence (as reflected by the increasing semi-variogram with time lag), and (iii) substantial monotone dropout with missingness increasing markedly at later visits. The fitted linear mixed-effects model is in broad agreement with these empirical features: the fixed effect of time is strongly positive (\(\hat\beta_{\text{time}}=7.22\), \(p<2\times10^{-16}\)), supporting an increasing mean structure, while the inclusion of subject-specific random intercepts and slopes (\(\widehat{\mathrm{Var}}(b_{0i})=1.443\), \(\widehat{\mathrm{Var}}(b_{1i})=0.685\)) captures between-subject heterogeneity and induces correlation among repeated measurements within a patient. Moreover, a random-slope specification implies a marginal variance that can increase with follow-up time, which is consistent with the qualitative pattern suggested by the smoothed empirical variance function. while the inclusion of baseline covariates (sex, education, job status, living situation, age, BMI, ADL, and baseline CDR-SB) captures the group-level differences observed in the LOESS plots. 

Exact agreement between exploratory summaries and the fitted mixed model is not required, and moderate differences do not automatically indicate model inadequacy. Empirical variance functions and nonparametric LOESS curves are descriptive tools that are sensitive to local data density; in particular, variance estimates at later visits are computed from a progressively smaller and more selected subset of participants due to dropout, making them inherently less stable and potentially affected by selection. In contrast, the fitted mixed model imposes a parsimonious parametric structure on both the mean and covariance: it summarizes the mean trajectory using a linear (fixed-effect) time trend and models within-subject dependence through random effects, while typically assuming a constant residual variance (here \(\widehat{\sigma}^2 = 2.697\)) unless time-varying heteroscedasticity is explicitly introduced. Consequently, the model is not expected to reproduce the raw empirical variance at each visit or match the LOESS smooths exactly. The relevant diagnostic criterion is whether the model captures the dominant mean trend and main covariate effects, provides an adequate description of the within-subject correlation structure, and yields acceptable residual behaviour for the inferential goals of the analysis.

## Sensitivity analysis for the continuous outcome (BPRS)

### Sensitivity analysis for the continuous outcome (BPRS) under MNAR
Given (i) increasing missingness over time and (ii) evidence that missingness relates to observed severity (baseline BPRS vs missed visits), sensitivity analysis focuses on plausible MNAR departures where unobserved post-dropout BPRS differs systematically from MAR-based predictions. 

To assess the potential impact of MNAR departures on substantive conclusions, a pattern-mixture \(\delta\)-adjustment sensitivity analysis was performed. First, the primary likelihood-based MAR linear mixed model for BPRS was fitted and used to generate MAR-based predictions \(\widehat{Y}^{MAR}_{ij}\) for all subject–visit combinations. For post-dropout occasions (\(j>t_i\)) with missing BPRS, MNAR-adjusted values were constructed as
$$
Y^{mis}_{ij}(\delta)=\widehat{Y}^{MAR}_{ij}+\delta,
$$
where \(\delta\) represents a systematic shift in the unobserved post-dropout outcomes relative to MAR predictions. The primary mixed model was then refitted over a grid of clinically plausible \(\delta\) values (negative \(\delta\) representing worse unobserved psychiatric status than predicted under MAR, and positive \(\delta\) the opposite), and the stability of key fixed-effect conclusions (notably the time effect and the time-by-baseline severity interaction) was evaluated. This provides a transparent “robustness curve” describing how strong MNAR departures would need to be to materially alter inference, acknowledging that MAR is not empirically verifiable from observed data alone.

This sensitivity analysis is anchored to the observed missingness–severity relationship in this cohort and provides a transparent way to quantify how strongly conclusions would need to depend on unobserved BPRS values to materially change inference. The need for such an analysis follows directly from the fact that the MAR assumption is not empirically verifiable. 



If you paste the *exact variable names* you use for (a) the time index and (b) the baseline covariate set you decided to carry forward (the ones you consistently used in your BPRS model), I can tighten the notation so the missingness model and $\delta$-sensitivity section match your final model specification perfectly (same $X_i$, same centering, same time coding).


```{r mnar-delta-sensitivity, message=FALSE, warning=FALSE}
library(data.table)
library(lme4)
library(lmerTest)

setDT(alzheimer_long)
setorder(alzheimer_long, patid, time)

# If time is 1..7, convert to 0..6
# alzheimer_long[, time := time - 1L]

# ---------------------------
# 1) Fit the primary MAR model
# ---------------------------
mar_fit <- lmer(
  bprs ~ time + age + sex + bmi + job + adl + wzc + cdrsb0 + time:cdrsb0 + (time | patid),
  data = alzheimer_long,
  REML = FALSE
)

# ---------------------------
# 2) Identify dropout time t_i (last observed BPRS time)
#    and mark post-dropout missing occasions (time > t_i)
# ---------------------------
alzheimer_long[, R_bprs := as.integer(!is.na(bprs))]

ti_dt <- alzheimer_long[R_bprs == 1, .(t_i = max(time)), by = patid]
alzheimer_long <- merge(alzheimer_long, ti_dt, by = "patid", all.x = TRUE)

# Post-dropout missingness indicator (for monotone dropout this captures all missing)
alzheimer_long[, post_dropout_missing := (is.na(bprs) & time > t_i)]

# ---------------------------
# 3) Get MAR-based predictions for all rows (subject-specific BLUP predictions)
# ---------------------------
alzheimer_long[, bprs_hat_mar := predict(mar_fit, newdata = alzheimer_long, re.form = NULL)]

# ---------------------------
# 4) Delta-adjustment grid and refit model under each delta
# ---------------------------
delta_grid <- c(-10, -5, -2, 0, 2, 5, 10)   # choose values meaningful on BPRS scale

fit_under_delta <- function(delta) {
  dt <- copy(alzheimer_long)

  # Impute ONLY post-dropout missing values: Y_mis(delta) = Yhat_MAR + delta
  dt[post_dropout_missing == TRUE, bprs := bprs_hat_mar + delta]

  # Refit the same model to the delta-imputed dataset
  fit <- lmer(
    bprs ~ time + age + sex + bmi + job + adl + wzc + cdrsb0 + time:cdrsb0 + (time | patid),
    data = dt,
    REML = FALSE
  )

  # Extract key fixed effects (edit terms if you want more)
  cf <- summary(fit)$coefficients
  terms_keep <- c("time", "time:cdrsb0")
  out <- data.table(
    delta = delta,
    term = terms_keep,
    estimate = cf[terms_keep, "Estimate"],
    se = cf[terms_keep, "Std. Error"],
    p_value = cf[terms_keep, "Pr(>|t|)"]
  )
  out
}

sens_results <- rbindlist(lapply(delta_grid, fit_under_delta))

sens_results

```


```{r}
library(ggplot2)

ggplot(sens_results, aes(x = delta, y = estimate, group = term)) +
  geom_line() + geom_point() +
  facet_wrap(~ term, scales = "free_y") +
  labs(title = "MNAR sensitivity (delta-adjustment): fixed-effect estimates vs delta",
       x = expression(delta), y = "Estimate") +
  theme_minimal()
```


### MNAR sensitivity analysis (pattern-mixture \(\delta\)-adjustment)

A \(\delta\)-adjustment sensitivity analysis was used to assess how departures from MAR could affect the main mixed-model conclusions for BPRS. Post-dropout missing BPRS values were shifted relative to their MAR-based predictions by \(Y^{mis}_{ij}(\delta)=\widehat{Y}^{MAR}_{ij}+\delta\), and the primary mixed model was refitted over a grid of \(\delta\in\{-10,-5,-2,0,2,5,10\}\). Across this range, the substantive conclusions were highly robust. The estimated time effect remained strongly positive and highly significant for all \(\delta\) values, varying smoothly from 6.22 per year at \(\delta=-10\) to 8.22 per year at \(\delta=10\) (all \(p<10^{-16}\)), indicating that the conclusion of increasing BPRS over time does not depend on the MAR assumption. Similarly, the time-by-baseline severity interaction remained negative and statistically significant throughout, ranging from \(-0.0155\) to \(-0.0285\) (all \(p \le 6.5\times10^{-4}\)), implying that higher baseline CDR-SB is consistently associated with a slower increase in BPRS over time. Overall, even under sizable MNAR shifts in post-dropout outcomes, the direction and statistical significance of the key fixed effects were preserved, suggesting that the main longitudinal inferences are not sensitive to plausible MNAR departures in this dataset.

