---
title: "LDA Homework 3 (2025-2026)"
author: 
  - Anita Kerubo Ogero (2469491)  
  - Okolie Tess Eucharia (2468507)  
  - Moses Mburu (2469245)  
  - Dorothy Chepkoech (2469284)

format:
  pdf:
    pdf-engine: xelatex
    fig-pos: "H"
    tbl-pos: "H"
    number-depth: 6
    include-in-header:
      text: |
        \renewcommand{\baselinestretch}{1.1}
        \usepackage{graphicx}
        \usepackage{xcolor}
        \usepackage{amsmath,amssymb}
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{array}
        \usepackage{multirow}
        \usepackage{makecell}
        \usepackage{pdflscape}
        \usepackage{float}
        \usepackage{fancyhdr}
  gfm: default

fontsize: 11pt
---





```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                        message = FALSE,
                        warning = FALSE,
                        fig.width = 7,
                        fig.height = 4.3)

library(kableExtra)

use_latex_packages()


tidy_with_ci <- function(model,
                         effects = c("fixed", "ran_pars"),
                         conf_level = 0.95,
                         exponentiate = FALSE,
                         digits = 2) {
  mm_class <- class(model)

  is_mixed <-
    any(c(
      "lmerMod", "lmerModLmerTest",
      "glmerMod", "nlmerMod"
    ) %in% mm_class) ||
      inherits(model, "glmmTMB")

  if (is_mixed) {
    if (!requireNamespace("broom.mixed", quietly = TRUE)) {
      stop("broom.mixed is required for mixed models.\n
        Install with install.packages('broom.mixed').")
    }
    out <- broom.mixed::tidy(
      model,
      effects = effects,
      conf.int = TRUE,
      conf.level = conf_level,
      exponentiate = exponentiate
    )
    # keep only common columns
    out <- out[, intersect(
      c(
        "effect", "group", "term", "estimate", "std.error",
        "statistic", "p.value", "conf.low", "conf.high"
      ),
      names(out)
    )]
  } else if (inherits(model, "gls")) {
    # nlme::gls objects are not supported by broom::tidy; extract manually
    s <- summary(model)
    if (!is.null(s$tTable)) {
      tt <- as.data.frame(s$tTable)
      # tTable columns: Value, Std.Error, t-value, p-value
      out <- data.frame(
        term = rownames(tt),
        estimate = tt[["Value"]],
        std.error = tt[["Std.Error"]],
        statistic = tt[["t-value"]],
        p.value = tt[["p-value"]],
        stringsAsFactors = FALSE
      )
      # Wald (normal) CIs
      z <- stats::qnorm(1 - (1 - conf_level) / 2)
      out$conf.low <- out$estimate - z * out$std.error
      out$conf.high <- out$estimate + z * out$std.error
      # ensure column names match later expectations
      out <- out[, c(
        "term", "estimate", "std.error",
        "statistic", "p.value", "conf.low", "conf.high"
      )]
    } else {
      stop("Unable to extract coefficient table from gls summary output.")
    }
  } else {
    if (!requireNamespace("broom", quietly = TRUE)) {
      stop("broom is required. Install with install.packages('broom').")
    }
    out <- broom::tidy(
      model,
      conf.int = TRUE,
      conf.level = conf_level,
      exponentiate = exponentiate
    )
    out <- out[, intersect(
      c(
        "term", "estimate", "std.error",
        "statistic", "p.value", "conf.low", "conf.high"
      ),
      names(out)
    )]
  }

  # formatted estimate with CI
  if (all(c("estimate", "conf.low", "conf.high") %in% names(out))) {
    # Identify rows that have valid CIs (i.e., not ran_pars)
    has_ci <- !is.na(out$conf.low)
    
    # Create the formatted estimate string, applying CI only where valid
    out$estimate <- ifelse(
      has_ci,
      paste0(
        sprintf(paste0("%.", digits, "f"), out$estimate), " [",
        sprintf(paste0("%.", digits, "f"), out$conf.low), ", ",
        sprintf(paste0("%.", digits, "f"), out$conf.high), "]"
      ),
      sprintf(paste0("%.", digits, "f"), out$estimate) # Just format the estimate for ran_pars
    )
  }

  # format numeric columns that are not the main estimate
  num_cols <- intersect(c("std.error", "statistic"), names(out))
  out[num_cols] <- lapply(out[num_cols], function(x) {
    if (is.numeric(x)) round(x, digits) else x
  })

  # optional: pretty p-values (e.g., <0.001)
  if ("p.value" %in% names(out)) {
    # Format p-values, leaving NAs as they are
    out$p.value <- ifelse(is.na(out$p.value), NA_character_,
      ifelse(out$p.value < 10^(-digits),
        paste0("<0.", paste(rep("0", digits - 1), collapse = ""), "1"),
        format(round(out$p.value, digits), nsmall = digits)
      )
    )
  }

  # reorder columns: put term first, then estimate string
  keep_order <- c(
    "effect", "group", "term", "estimate", "std.error",
    "statistic", "p.value"
  )
  out <- out[, intersect(keep_order, names(out))]

  tibble::as_tibble(out)
}



# Tidy adjacent model comparisons from base anova() output
# Works for anova(m1, m2, ...): lm/glm, lmer/glmer, gls/lme, etc.
tidy_anova_compare <- function(anova_df, digits = 2) {
  x <- anova_df
  is_num <- vapply(x, is.numeric, logical(1))
  x[is_num] <- lapply(x[is_num], function(v) {
    ifelse(is.na(v), v, round(v, digits))
  })
  x$call <- NULL
  x
}



display_random_effects <- function(
    model,
    caption = "Random effects: variance components",
    digits = 3) {
  df <- tidy_re_varcorr(model, digits = digits)
  knitr::kable(
    df,
    format = "latex",
    booktabs = TRUE,
    #caption = caption,
    align = c("l", "l", "r", "r", "r")
  ) |>
    kableExtra::kable_styling(latex_options = "hold_position")
}



tidy_re_varcorr <- function(model, digits = 3) {
  vc <- lme4::VarCorr(model)
  rows <- list()

  for (grp in names(vc)) {
    m <- vc[[grp]]
    sd <- attr(m, "stddev")
    trm <- names(sd)
    var <- sd^2

    corMat <- tryCatch(attr(m, "correlation"),
      error = function(e) NULL
    )

    for (k in seq_along(trm)) {
      term <- trm[k]
      corr_val <- NA_real_
      if (!is.null(corMat) && nrow(corMat) > 1 &&
        term != "(Intercept)" &&
        "(Intercept)" %in% colnames(corMat)) {
        corr_val <- as.numeric(corMat[term, "(Intercept)"])
      }
      rows[[length(rows) + 1L]] <- data.frame(
        Groups = grp,
        Name = term,
        Variance = var[k],
        Std.Dev. = sd[k],
        Corr = corr_val,
        check.names = FALSE
      )
    }
  }

  out <- do.call(rbind, rows)

  # Add Residual variance
  resid_sd <- sigma(model)
  resid_var <- resid_sd^2
  resid_row <- data.frame(
    Groups = "Residual",
    Name = "",
    Variance = resid_var,
    Std.Dev. = resid_sd,
    Corr = NA_real_,
    check.names = FALSE
  )
  out <- rbind(out, resid_row)

  num_cols <- c("Variance", "Std.Dev.", "Corr")
  out[num_cols] <- lapply(
    out[num_cols],
    function(x) {
      ifelse(is.na(x),
        NA, round(x, digits)
      )
    }
  )
  out
}


```



```{r}
library(here)
library(tidyverse)
library(data.table)
library(haven)
library(janitor)
library(knitr)
library(lme4)   
library(lmerTest)
library(pbkrtest)
library(gridExtra)

lda_path <- here()
lda_data_path <- here(lda_path, "data")
#list.files(lda_data_path)

```



```{r}
# Load the data
alzheimer_data <- read_sas(here(
    lda_data_path,
    "alzheimer25.sas7bdat"
)) |> clean_names()


```





```{r factor-conversion}
setDT(alzheimer_data)
# log bmi
#alzheimer_data[, bmi := log(bmi)]
factor_info <- list(
  sex = list(
    levels = c(0, 1),
    labels = c("Male", "Female")
  ),
  edu = list(
    levels = c(1, 2, 3, 4),
    labels = c("Primary", "Lower secondary", "Upper secondary", "Higher")
  ),
  job = list(
    levels = c(0, 1),
    labels = c("No Paid Job", "Paid Job")
  ),
  wzc = list(
    levels = c(0, 1),
    labels = c("Lives at Home", "Nursing Home")
  )
)


for (col_name in names(factor_info)) {
  info <- factor_info[[col_name]]
  alzheimer_data[, (col_name) := factor(get(col_name),
    levels = info$levels,
    labels = info$labels
  )]
}

# Verify the changes by checking the structure of the new columns
#kable(alzheimer_data[1:5, c("sex", "edu", "job", "wzc")])
```


```{r reshape-data}

id_vars <- c("patid", "trial", "sex", "age", 
             "edu", "bmi", "inkomen", "job", "adl", "wzc",
            "cdrsb0",  "abpet0", "taupet0" )



alzheimer_long_cdrsb <- melt(alzheimer_data,
  id.vars = id_vars,
  measure.vars = patterns(
    cdrsb = "^cdrsb[0-6]",
    bprs = "^bprs[0-6]",
    abpet = "^abpet[0-6]",
    taupet = "^taupet[0-6]"
  ),
  variable.name = "time",
  value.name = c("cdrsb", "bprs", "abpet", "taupet"),
  variable.factor = FALSE
)

alzheimer_long_cdrsb[, time := as.integer(time)]

#alzheimer_long[, time := time - 1]


# Display the first few rows of the new long-format data
#kable(head(alzheimer_long))



alzheimer_long_bprs <- melt(alzheimer_data,
  id.vars = id_vars,
  measure.vars = patterns(
    cdrsb = "^cdrsb[1-6]",
    bprs = "^bprs[0-6]",
    abpet = "^abpet[1-6]",
    taupet = "^taupet[1-6]"
  ),
  variable.name = "time",
  value.name = c("cdrsb", "bprs", "abpet", "taupet"),
  variable.factor = FALSE
)

alzheimer_long_bprs[, time := as.integer(time)]
```






## Methodology

### Notation and data structure

Let $i=1,\dots,N$ index patients and $j=0,\dots,J$ index scheduled annual visits ($j=0$ baseline). The continuous outcome of interest is the BPRS score $Y_{ij}$. Define the missingness indicator

$$
R_{ij} =
\begin{cases}
1, & \text{if } Y_{ij}\ \text{is observed},\
0, & \text{if } Y_{ij}\ \text{is missing.}
\end{cases}
$$

Let $R_i=(R_{i0},\dots,R_{iJ})$ denote the subject-specific observation pattern, and split the full outcome vector as $Y_i=(Y_i^{obs},Y_i^{mis})$.



### Missingness pattern and dropout taxonomy

The first goal is to describe whether missingness behaves like monotone dropout (typical in longitudinal follow-up) or intermittent missingness (temporary gaps). For each patient:

1. **Completeness.** A patient is complete if $R_{ij}=1$ for all $j$.

2. **Monotone missingness (dropout-type).** Define the last observed visit
   $$
   t_i=\max{j: R_{ij}=1}.
   $$


   The pattern is monotone if


   $$
   R_{ij}=1 \text{ for } j\le t_i,\qquad R_{ij}=0 \text{ for } j>t_i.
   $$


3. **Non-monotone (intermittent) missingness.** The pattern is non-monotone if there exist $j<k$ such that
   $$
   R_{ij}=0 \ \text{and}\ R_{ik}=1.
   $$

These classifications are summarised using (i) the distribution of $t_i$ (retention over time), (ii) the proportion missing at each visit $j$, and (iii) the frequency of intermittent gaps within $R_i$. In this Alzheimer cohort, missingness rises markedly with follow-up, so later-visit summaries are based on a progressively smaller subset of participants. 


###  Proportion missing by time

```{r}

library(data.table)
library(ggplot2)

# Ensure data.table
setDT(alzheimer_long)
setorder(alzheimer_long, patid, time)

alzheimer_long[, time := time - 1L]


# --- Choose outcome to study (start with BPRS)
outcome <- "bprs"

# 1) Create missingness indicator R_ij for chosen outcome
alzheimer_long[, R := as.integer(!is.na(get(outcome)))]

# 2) Proportion missing by time
miss_by_time <- alzheimer_long[, .(
  n_total = .N,
  n_obs   = sum(R == 1),
  n_miss  = sum(R == 0),
  prop_miss = mean(R == 0)
), by = time][order(time)]

miss_by_time
```




```{r}
ggplot(miss_by_time, aes(x = time, y = prop_miss)) +
  geom_line(color = "blue", size = 1) + geom_point(color = "blue", size = 2) +
  scale_x_continuous(breaks = 0:6) +
  labs(title = paste0("Proportion missing in ", outcome, " by visit"),
       x = "Time (0=baseline, ..., 6)", y = "Proportion missing") +
  theme_minimal()
```


### Subject-level pattern features: t_i, monotone vs non-monotone, gaps

```{r}
# 3) Subject-level pattern features: t_i, monotone vs non-monotone, gaps
# Helper: monotone dropout if once missing => always missing
is_monotone <- function(Rvec) all(cummin(Rvec) == Rvec)

pattern_dt <- alzheimer_long[, {
  Rvec <- R
  # last observed time (t_i); if all missing, set NA (rare if baseline always observed)
  if (all(Rvec == 0)) {
    list(t_i = NA_integer_, complete = FALSE, monotone = TRUE, non_monotone = FALSE, n_gaps = NA_integer_)
  } else {
    t_i <- max(time[Rvec == 1])
    n_gaps <- sum(Rvec == 0 & time < t_i)   # missing before last observed => intermittent gap(s)
    mono <- is_monotone(Rvec)
    list(
      t_i = t_i,
      complete = all(Rvec == 1),
      monotone = mono,
      non_monotone = !mono,
      n_gaps = n_gaps
    )
  }
}, by = patid]

pattern_dt[, pattern := fifelse(complete, "Complete",
                        fifelse(monotone, "Monotone dropout", "Non-monotone (intermittent)"))]

# Pattern counts
pattern_dt[, .N, by = pattern]

```


###Distribution of t_i (retention / dropout timing)

```{r}
# 4) Distribution of t_i (retention / dropout timing)
ti_dist <- pattern_dt[!is.na(t_i), .N, by = t_i][order(t_i)]
ti_dist

ggplot(pattern_dt[!is.na(t_i)], aes(x = factor(t_i))) +
  geom_bar(fill = "blue") +
  labs(title = paste0("Dropout timing: distribution of t_i (", outcome, ")"),
       x = "t_i (last observed visit)", y = "Number of patients") +
  theme_minimal()
```



### Intermittent missingness summary
```{r}
# 5) Intermittent missingness summary
pattern_dt[, .(
  n_patients = .N,
  n_non_monotone = sum(pattern == "Non-monotone (intermittent)"),
  prop_non_monotone = mean(pattern == "Non-monotone (intermittent)"),
  mean_gaps_non_mon = mean(n_gaps[pattern == "Non-monotone (intermittent)"], na.rm = TRUE)
)]
```


```{r}
# 6) Show example R_i sequences for a few non-monotone patients
example_ids <- pattern_dt[pattern == "Non-monotone (intermittent)"][1:5, patid]

Ri_examples <- alzheimer_long[patid %in% example_ids,
                              dcast(.SD, patid ~ time, value.var = "R")]

Ri_examples
```

### Missingness Conclusion

Missingness in BPRS increases markedly with follow-up, reaching its highest level at the final visit, indicating substantial attrition over time. Examination of subject-specific observation patterns shows that missingness is entirely monotone: participants who miss a BPRS measurement do not re-enter the study at later visits. Specifically, 742 participants exhibit monotone dropout and 511 are complete cases; no intermittent (non-monotone) missingness is observed. The distribution of the last observed visit $t_i$ confirms dropout occurs throughout follow-up, so later-visit summaries represent a progressively selected subset of patients.

Because follow-up is incomplete, the data are unbalanced, with progressively fewer observations at later visits. This makes methods that require complete data (such as repeated-measures ANOVA) inappropriate and potentially biased. The missingness pattern is monotone, consistent with dropout, so it is reasonable to work under a MAR framework (at least as a starting assumption) and use models that properly accommodate dropout and within-subject dependence. In this setting, likelihood-based mixed-effects models are recommended such as linear mixed models for the continuous outcome (BPRS) and GLMMs for the binary outcome because they can use all available data despite unequal numbers of measurements per patient, explicitly account for within-patient correlation, and provide valid inference under MAR when the mean and covariance structures are correctly specified.


### Empirical signals of informative missingness

Because BPRS is measured repeatedly, a second goal is to evaluate whether missingness is associated with observed disease severity/history (evidence against MCAR and supportive of MAR as a working assumption). Two descriptive comparisons are used:

* **Baseline severity vs missingness burden.** Let $M_i=\sum_{j=1}^J (1-R_{ij})$ be the number of missed follow-up BPRS measurements. The relationship between baseline BPRS ($Y_{i0}$) and $M_i$ is examined to assess whether those with worse baseline psychiatric symptoms are more likely to miss subsequent visits.

Figure X shows the relationship between baseline BPRS ($Y_i0$) and the number of missed follow-up visits ($M_i$). There is a clear positive association: patients with higher baseline BPRS scores tend to miss more subsequent follow-up measurements. In particular, individuals with low baseline BPRS are predominantly complete or have few missed visits, whereas those with higher baseline psychiatric symptom severity exhibit substantially greater missingness burden. This pattern indicates that dropout is related to observed disease severity, providing strong evidence against a MCAR mechanism and supporting MAR as a reasonable working assumption for the missingness process. Because missingness depends on observed baseline severity, likelihood-based mixed-effects models that are valid under MAR are appropriate for primary longitudinal analyses.

```{r baseline-severity-missingness-burden}
library(data.table)
library(ggplot2)

setDT(alzheimer_long)
setorder(alzheimer_long, patid, time)

# Ensure time is 0..6 (if yours is 1..7, uncomment next line)
# alzheimer_long[, time := time - 1L]

# Missingness indicator for BPRS
alzheimer_long[, R_bprs := as.integer(!is.na(bprs))]

# Subject-level: baseline BPRS and number of missed follow-ups (exclude baseline j=0)
burden_dt <- alzheimer_long[, .(
  bprs0 = bprs[time == 0][1],
  M_i   = sum(1 - R_bprs[time >= 1])  # missed follow-ups only
), by = patid]

# Quick summary table
burden_dt[, .(
  n = .N,
  mean_M = mean(M_i, na.rm = TRUE),
  sd_M   = sd(M_i, na.rm = TRUE),
  min_M  = min(M_i, na.rm = TRUE),
  max_M  = max(M_i, na.rm = TRUE)
)]

# Scatter + smooth trend: baseline severity vs missingness burden
ggplot(burden_dt, aes(x = bprs0, y = M_i)) +
  geom_jitter(color = "blue", width = 0.2, height = 0.1, alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Baseline BPRS vs missing follow-up burden (M_i)",
       x = "Baseline BPRS (Y_i0)", y = "Missed follow-up visits (M_i)") +
  theme_minimal()

```


* **Visit-specific missingness rates.** The proportion missing at each visit $j$,

  $$
  \widehat{p}*j=\frac{1}{N}\sum*{i=1}^N (1-R_{ij}),
  $$

  is plotted over time to characterise the growth of missingness with follow-up (and, indirectly, the degree of selection in later years). 
  
Figure Y displays the visit-specific missingness rates $𝑝^𝑗$ for BPRS over time. Missingness increases monotonically with follow-up, from essentially zero at baseline to approximately 60% by the final visit. This steady rise indicates substantial attrition and increasing selection among participants who remain under observation at later time points. Consequently, summaries at later visits are based on progressively smaller and more selected subsets of patients, reinforcing the need for modeling approaches that can accommodate unbalanced longitudinal data and informative dropout.

The strong time dependence of $𝑝^𝑗$ further rules out MCAR and motivates likelihood-based mixed-effects models that remain valid under MAR.

These summaries establish whether the data resemble monotone dropout, whether intermittent missingness occurs, and whether observed severity/history is linked to being missing—key inputs for choosing MAR-valid modelling strategies later.



```{r visit-specific-missingness-rates}
# Visit-specific missingness for BPRS
miss_by_visit <- alzheimer_long[, .(
  N = uniqueN(patid),
  n_obs = sum(R_bprs == 1),
  n_miss = sum(R_bprs == 0),
  p_hat = mean(R_bprs == 0)
), by = time][order(time)]

miss_by_visit

ggplot(miss_by_visit, aes(x = time, y = p_hat)) +
  geom_line(color = "blue") + geom_point(color = "blue") +
  scale_x_continuous(breaks = 0:6) +
  labs(title = "Visit-specific missingness rate for BPRS",
       x = "Visit (time)", y = expression(hat(p)[j])) +
  theme_minimal()
```


### Working assessment of MCAR vs MAR, and why MNAR cannot be “diagnosed” from observed data

To formalise the dependence of missingness on observed history, a discrete-time observation model is fitted. Define a “next-visit missingness” indicator for those observed at visit $j$:

$$
D_{i,j+1} = 1-R_{i,j+1}.
$$

A logistic regression (discrete-time hazard) model is used:

$$
\Pr(D_{i,j+1}=1 \mid \mathcal{H}*{ij})
= \operatorname{logit}^{-1}!\Big(
\alpha_0 + \alpha_1 Y*{ij} + \alpha_2^\top X_i + \alpha_3 j
\Big),
$$


where $\mathcal{H}*{ij}$ is the observed history up to $j$ and $X_i$ contains baseline covariates (e.g., sex, age, BMI, ADL, job status, living situation, baseline CDR-SB, and other baseline markers as available). A clear dependence of $D*{i,j+1}$ on observed outcomes/covariates argues against MCAR and supports MAR as a reasonable primary assumption.

The MAR mechanism is stated as

$$
p(R_i \mid Y_i, X_i)=p(R_i \mid Y_i^{obs}, X_i),
$$

whereas MNAR allows dependence on $Y_i^{mis}$. Since MNAR is defined through unobserved outcomes, it cannot be identified from the observed data alone; therefore, departures from MAR are addressed through sensitivity analysis rather than “testing” MNAR.



```{r discrete-time-dropout-model}
library(data.table)

setDT(alzheimer_long)
setorder(alzheimer_long, patid, time)

# Ensure time is 0..6 (if yours is 1..7, uncomment next line)
# alzheimer_long[, time := time - 1L]

# 1) Missingness indicator for current visit (R_ij)
alzheimer_long[, R := as.integer(!is.na(bprs))]

# 2) Construct next-visit missingness indicator:
#    D_{i,j+1} = 1 - R_{i,j+1}
alzheimer_long[, D_next := 1L - shift(R, type = "lead"), by = patid]

# 3) Keep only rows where:
#    - current visit is observed (R_ij = 1), so the "risk set" at time j
#    - next visit exists (time <= 5 if visits are 0..6)
obs_model_dt <- alzheimer_long[
  R == 1 & time <= 5,
  .(patid, time, D_next,
    y_ij = bprs,                 # Y_ij = current observed BPRS
    age, sex, bmi, job, adl, wzc,
    cdrsb0, abpet0, taupet0, edu)
]

# Optional: drop rows where D_next is NA (should only happen at last visit)
obs_model_dt <- obs_model_dt[!is.na(D_next)]

# 4) Fit discrete-time hazard (logistic regression)
dropout_glm <- glm(
  D_next ~ y_ij + age + sex + bmi + job + adl + wzc + cdrsb0 + abpet0 + taupet0 + edu + time,
  data = obs_model_dt,
  family = binomial
)

summary(dropout_glm)

# Optional: odds ratios (more interpretable)
OR <- exp(cbind(OR = coef(dropout_glm), confint(dropout_glm)))
OR

```

The discrete-time observation (dropout) model provides strong evidence that missingness depends on observed history and observed covariates, thereby ruling out MCAR as a plausible mechanism. The probability of being missing at the next visit increases sharply with follow-up time (odds ratio \(=3.27\) per year, \(p<2\times10^{-13}\)), indicating substantial time-dependent attrition. Several baseline clinical characteristics are also strongly associated with next-visit missingness: ADL is highly predictive (odds ratio \(=0.15\), \(p<2\times10^{-16}\)), nursing home residence markedly increases the risk of dropout (odds ratio \(=6.05\), \(p<2\times10^{-16}\)), and baseline CDR-SB shows a significant positive association with dropout (odds ratio \(=1.05\) per unit, \(p=4.3\times10^{-7}\)). 

In contrast, the current observed BPRS level \(Y_{ij}\) is not statistically significant (\(p=0.15\)), although its estimated effect is in the positive direction. Overall, because missingness clearly depends on observed covariates and visit time, the data are inconsistent with MCAR and support MAR as a reasonable primary working assumption.

Although MAR is supported by the observed data, an MNAR mechanism—where missingness depends on unobserved future outcomes cannot be ruled out. By definition, MNAR involves dependence on unobserved responses and therefore cannot be diagnosed using the observed data alone; potential departures from MAR must instead be investigated through sensitivity analyses.


### MAR assumption used for primary analyses (and link to later models)

For primary modelling, missingness is treated as ignorable under MAR (with distinct parameters for the outcome and missingness processes), so inference can proceed using the observed-data likelihood for the continuous outcome models. This aligns with the modelling assumptions used for BPRS longitudinal mixed modelling in this dataset. 

In addition, the fitted observation model above yields estimated observation probabilities
$$
\widehat{\pi}*{ij}=\Pr(R*{ij}=1 \mid \mathcal{H}*{i,j-1}),
$$

which can be converted into inverse-probability weights $w*{ij}=1/\widehat{\pi}_{ij}$ for MAR-valid weighted marginal analyses when needed.



# Question2. In the first assignment, linear mixed models have been considered. Select one of them

```{r}
library(lme4 )
fm_adj <- bprs ~ time + sex + 
  age + bmi + job + adl + wzc +  cdrsb0 + 
   time:cdrsb0 +
   1 + (time | patid) + (1 | trial)
lmer_reduced_mean_structure <- lmer(fm_adj,
  data = alzheimer_long,
  na.action = na.exclude
)

summary(lmer_reduced_mean_structure)
```



### Question 3 For the binary outcome, consider a random-effects as well as a marginal model.

#### Marginal Model

```{r reshape-data}

# id_vars <- c("patid", "trial", "sex", "age", 
# "edu", "bmi", "inkomen", "job", "adl", "wzc")
          



alzheimer_long <- melt(alzheimer_data,
  id.vars = id_vars,
  measure.vars = patterns(
    cdrsb = "^cdrsb[0-6]",
    bprs = "^bprs[0-6]",
    abpet = "^abpet[0-6]",
    taupet = "^taupet[0-6]"
  ),
  variable.name = "time",
  value.name = c("cdrsb", "bprs", "abpet", "taupet"),
  variable.factor = FALSE
)

alzheimer_long[, time := as.integer(time)]

alzheimer_long[, CDRSBbin := ifelse(cdrsb > 10, 1, 0)]
#alzheimer_long[, time := time - 1]


# Display the first few rows of the new long-format data
#kable(head(alzheimer_long))
```



```{r}
library(geepack)
# Ensure data is sorted for GEE
setorder(alzheimer_long, patid, time)

gee_model <- geeglm(
    CDRSBbin ~ time + sex + age + bmi +
     adl + abpet + taupet + 
     time:abpet + time:taupet,
    id = patid,
    data = alzheimer_long,
    family = binomial(link = "logit"),
    corstr = "ar1",
    std.err = "san.se" # Robust standard errors
)

summary(gee_model)
```



#### Random Effects model

```{r}
alzheimer_long <- alzheimer_long |>
  dplyr::mutate(
    time_c   = time   - mean(time,   na.rm = TRUE),
    age_c    = age    - mean(age,    na.rm = TRUE),
    bmi_c    = bmi    - mean(bmi,    na.rm = TRUE),
    adl_c    = adl    - mean(adl,    na.rm = TRUE),
    abpet_c  = abpet  - mean(abpet,  na.rm = TRUE),
    taupet_c = taupet - mean(taupet, na.rm = TRUE)
  )

ctrl <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
glmm_center <- glmer(
  CDRSBbin ~ time_c + sex + age_c + bmi_c + adl_c +
    abpet_c + taupet_c + time_c:taupet_c +
    (time_c | patid),
  data   = alzheimer_long,
  family = binomial(link = "logit"),
  nAGQ   = 1,
  control = ctrl
)
summary(glmm_center)
```


### Question 3  For each of these three models, present one version that is valid under MAR. While thereare several possible solutions in each of the three cases, it is explicitly the idea to select one of them and then argue why you have made this choice


1. Continuous outcome (BPRS): MAR-valid Linear Mixed Model (likelihood-based)
MAR is valide becauseLikelihood-based LMM uses the observed-data likelihood and is valid under MAR (with correct mean/covariance).

```{r mar-lmm-bprs}
library(data.table)
library(lme4)
library(lmerTest)

setDT(alzheimer_long)
setorder(alzheimer_long, patid, time)

# Ensure time is 0..6 (uncomment if yours is 1..7)
# alzheimer_long[, time := time - 1L]

# (adjust covariates to match your final selection)
lmm_mar <- lmer(
  bprs ~ time + age + sex + bmi + job + adl + wzc + cdrsb0 + time:cdrsb0 + (time | patid),
  data = alzheimer_long,
  REML = FALSE
)

summary(lmm_mar)
```

2. Binary outcome (CDRSBbin): MAR-valid Random-effects logistic model (GLMM).MAR-valid Because Likelihood-based GLMM is valid under MAR (again, under correct model + ignorability)

```{r mar-glmm-cdrsbbin}
library(lme4)

# Make sure CDRSBbin exists
alzheimer_long[, CDRSBbin := ifelse(cdrsb > 10, 1L, 0L)]

# Random intercept logistic model (you can use (time|patid) if needed)
glmm_mar <- glmer(
  CDRSBbin ~ time_c + sex + age_c + bmi_c + adl_c +
    abpet_c + taupet_c + time_c:taupet_c +
    (time_c | patid),
  data = alzheimer_long,
  family = binomial,
  nAGQ = 1
)

summary(glmm_mar)
```

3. Weighted GEE
### Problem: why standard GEE is not generally MAR-valid

Standard GEE yields consistent estimation under **MCAR** (and in some special cases), but it is **not generally valid under MAR** when dropout depends on observed history. In our data, missingness increases with follow-up and is associated with observed severity and covariates, so a MAR-valid marginal approach is required.

### MAR-valid version: Weighted GEE (IPW-GEE)

To obtain a marginal model that is valid under MAR, we use **inverse-probability weighted GEE (IPW-GEE)**. First, an observation (missingness) model is fitted to estimate the probability that subject \(i\) is observed at visit \(j\), given the observed history up to the previous visit:

$$
\widehat{\pi}_{ij} = \Pr\!\left(R_{ij}=1 \mid \mathcal{H}_{i,j-1}\right).
$$

Inverse probability weights are then constructed as:

$$
w_{ij}=\frac{1}{\widehat{\pi}_{ij}}.
$$

Next, the marginal mean model for the binary outcome is fitted:

$$
\Pr(Y_{ij}=1)
= \operatorname{logit}^{-1}\!\left(\beta_0+\beta_1\,\text{time}_{ij}+\beta^\top X_i\right),
$$

using GEE with weights \(w_{ij}\).

### Why IPW-GEE is MAR-valid

If the observation model for \(\pi_{ij}\) is correctly specified, the inverse-probability weights correct for the selection induced by MAR dropout. Therefore, the weighted estimating equations yield **consistent marginal (population-averaged) estimates under MAR**.

**Justification (choice of method):**

I use IPW-weighted GEE because standard GEE is not generally valid under MAR dropout; inverse probability weights based on the fitted observation model yield consistent marginal estimates under MAR.


```{r}
# ============================================================
# MAR-valid Weighted GEE (IPW-GEE) for CDRSBbin
#   - Step 1: fit a dropout/observation model (discrete-time)
#   - Step 2: build inverse-probability-of-observation weights
#   - Step 3: fit weighted GEE (geeglm)
# ============================================================

library(data.table)
library(geepack)

setDT(alzheimer_long)
setorder(alzheimer_long, patid, time)

# If your time is 1..7, convert to 0..6
# alzheimer_long[, time := time - 1L]

# -----------------------------
# Outcome + missingness indicator
# -----------------------------
alzheimer_long[, CDRSBbin := ifelse(!is.na(cdrsb) & cdrsb > 10, 1L,
                            ifelse(!is.na(cdrsb), 0L, NA_integer_))]

# Observed indicator for continuous cdrsb (same missingness as CDRSBbin)
alzheimer_long[, R := as.integer(!is.na(cdrsb))]

# -----------------------------
# Step 1: Discrete-time dropout model
# D_{i,j+1} = 1 - R_{i,j+1}, among those observed at time j
# -----------------------------
alzheimer_long[, D_next := 1L - shift(R, type = "lead"), by = patid]

# Risk set: observed at time j, and j+1 exists (if time 0..6, last j is 5)
obs_dt <- alzheimer_long[R == 1 & time <= 5,
                         .(patid, time, D_next,
                           y_ij = cdrsb,   # observed history at time j
                           age, sex, bmi, job, adl, wzc, cdrsb0, edu)]

obs_dt <- obs_dt[!is.na(D_next)]

# Fit dropout model (you can drop covariates that cause separation if needed, e.g., job)
dropout_glm <- glm(
  D_next ~ time + y_ij + age + sex + bmi + job + adl + wzc + cdrsb0 + edu,
  data = obs_dt,
  family = binomial
)

# Predict conditional probability of being observed at next visit:
# p_obs_next(i,j) = P(R_{i,j+1}=1 | history up to j, R_{ij}=1)
obs_dt[, p_drop_next := predict(dropout_glm, type = "response")]
obs_dt[, p_obs_next  := pmax(1 - p_drop_next, 1e-6)]  # guard against 0

# Create table mapping p_obs_next to the NEXT visit (time+1)
p_next <- obs_dt[, .(patid, time = time + 1L, p_obs_step = p_obs_next)]

# -----------------------------
# Step 2: Build cumulative observation probabilities pi_it
# pi_it = Π_{k=1..t} p_obs_step(i,k)
# where p_obs_step(i,k) = P(observed at time k | observed at time k-1, history)
# -----------------------------
w_dt <- unique(alzheimer_long[, .(patid, time)])
setorder(w_dt, patid, time)

# Merge step-probabilities onto each (patid, time)
w_dt <- merge(w_dt, p_next, by = c("patid", "time"), all.x = TRUE)

# Baseline time=0: set step prob = 1 (always observed at baseline in your study)
w_dt[time == 0, p_obs_step := 1]

# For visits with no model-based probability (should be only impossible cases), set NA to 1
w_dt[is.na(p_obs_step), p_obs_step := 1]

# Cumulative probability of being observed up to time t
w_dt[, pi_obs := cumprod(p_obs_step), by = patid]

# Stabilized weights (recommended): numerator = marginal P(observed at time)
# Compute marginal observation rate by time:
marg_obs <- alzheimer_long[, .(p_marg_obs = mean(R == 1, na.rm = TRUE)), by = time]
w_dt <- merge(w_dt, marg_obs, by = "time", all.x = TRUE)
w_dt[is.na(p_marg_obs), p_marg_obs := 1]

# Stabilized cumulative numerator
w_dt[, num_pi := cumprod(p_marg_obs), by = patid]

# Stabilized IPW
w_dt[, w_ipw := num_pi / pi_obs]

# Optional: truncate extreme weights for stability (common)
cap <- quantile(w_dt$w_ipw, 0.99, na.rm = TRUE)
w_dt[w_ipw > cap, w_ipw := cap]

# Attach weights back to main long data
alzheimer_long <- merge(alzheimer_long, w_dt[, .(patid, time, w_ipw)], by = c("patid", "time"), all.x = TRUE)

# -----------------------------
# Step 3: Weighted GEE (marginal model) for observed CDRSBbin
# -----------------------------
gee_dt <- alzheimer_long[!is.na(CDRSBbin)]

wgee_mar <- geeglm(
  CDRSBbin ~ time + age + sex + bmi + job + adl + wzc + cdrsb0 + edu,
  id = patid,
  data = gee_dt,
  family = binomial,
  corstr = "exchangeable",
  weights = w_ipw
)

summary(wgee_mar)

```


### Question 5 Examine whether the results obtained from data exploration in the first assignment arein agreement with the results obtained from some of these models. As an example: isthe empirical variance function of Assignment 1 sufficiently similar to the fitted variancefunction in the version of the linear mixed model considered here? Would it be a problem if they differ, and why (not)?

```{r}
#| label: fig-missingness-analysis
#| fig-cap: "Missingness Analysis: Baseline BPRS by Number of Missed Visits (left) and Percentage Missing Over Time (right)"
#| fig.width: 8.5
#| fig.height: 3.5

bprs_cols <- paste0("bprs", 0:6)
alzheimer_data[, n_missing := rowSums(is.na(.SD)), .SDcols = bprs_cols]

# Group by the number of missing visits
# and calculate summary stats for baseline BPRS
dropout_summary <- alzheimer_data[!is.na(bprs0), .(
  mean_bprs0 = mean(bprs0),
  median_bprs0 = median(bprs0),
  visit_n_patients = .N
), by = n_missing][order(n_missing)]



missing_by_time <- alzheimer_long[, .(
  n_patients = uniqueN(patid),
  n_obs = sum(!is.na(bprs))
), by = time][order(time)]

missing_by_time[, pct_missing := round(100 *
                                         (n_patients - n_obs)/
                                         n_patients, 1)]



boxplot_meanbprso0 <- ggplot(alzheimer_data[!is.na(bprs0)], 
                             aes(x = factor(n_missing),
                                 y = bprs0)) +
  geom_boxplot(aes(fill = factor(n_missing)),
               alpha = 0.7) +
  labs(
    title = "Baseline BPRS Score by
    Number of Missed Visits",
    x = "Number of Missed 
    BPRS Measurements",
    y = "Baseline BPRS 
    Score (bprs0)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

missing_by_time_plot <- ggplot(missing_by_time, 
                               aes(x = time, y = pct_missing)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue", size = 2) +
  labs(
    title = "Percentage of Missing BPRS \n Measurements Over Time",
    x = "Time (Years)",
    y = "Percentage of Missing BPRS Measurements (%)"
  ) +
  theme_minimal()

grid.arrange(boxplot_meanbprso0,
             missing_by_time_plot,
             ncol = 2
             )
```

```{r}
#| label: loess-smoothing
#  LOESS model

loess_fit <- loess(bprs ~ time,
  data = alzheimer_long[!is.na(bprs)],
  na.action = na.omit
)

# Score the data to get predictions (equivalent to `score data=...`)

bprs_smoothed <- predict(loess_fit, newdata = alzheimer_long)

# Create and sort the output data

loess_results <- alzheimer_long[!is.na(bprs), .(patid, time, bprs, sex)]
loess_results[, predicted_bprs := predict(loess_fit, newdata = .SD)]

# Sort the results by time
setorder(loess_results, time)


# Visualize 
simple_avg_plot <- ggplot(alzheimer_long, aes(x = time, y = bprs)) +
  geom_point(alpha = 0.1, color = "gray") + # Plot raw data points transparently
  geom_line(
    data = loess_results,
    aes(y = predicted_bprs), color = "blue", size = 1
  ) +
  labs(
    title = "LOESS Smoothed Trend of BPRS Over Time",
    x = "Time (Years)",
    y = "BPRS Score"
  ) +
  theme_minimal()

simple_avg_plot
```


```{r}
#| label: fig-loess-by-categorical-vars
#| fig-width: 10.3
#| fig-height: 8
#| fig-cap: "LOESS Smoothed BPRS Over Time by Subgroups. Overall average (top),  sex, education, job status, living situation, and clinical center."


library(gridExtra)

subgroup_vars <- c("sex", 
                   "edu", 
                   "job",
                   "wzc", 
                   "trial")

# factor trial

alzheimer_long[, trial := factor(as.character(trial))]

mycolors <- c(
  "#E41A1C", "#377EB8", "#4DAF4A",
  "#984EA3", "#FF7F00", "#FFFF33",
  "#A65628", "#1B9E77", "#D95F02",
  "#7570B3", "#E7298A", "#66A61E",
  "#E6AB02", "#A6761D", "#7FC97F",
  "#BEAED4", "#FDC086", "#FFFF99",
  "#386CB0", "#F0027F", "#BF5B17",
  "#FBB4AE", "#B3CDE3", "#CCEBC5",
  "#DECBE4", "#FED9A6", "#FFFFCC", 
  "#E5D8BD"
)
# Titles for each plot
plot_titles <- c(
  "sex" = "LOESS Smoothed BPRS by Sex",
  "edu" = "LOESS Smoothed BPRS by Education",
  "job" = "LOESS Smoothed BPRS by Job Status",
  "wzc" = "LOESS Smoothed BPRS by Living Situation",
  "trial" = "LOESS Smoothed BPRS by Clinical Center",
  "taupet0" = "LOESS Smoothed BPRS by Baseline Tau PET"
)

#alzheimer_long[, taupet0 := factor(as.character(taupet0))]
# List to store plots
plot_list <- list()

# Loop to create plots
for (var in subgroup_vars) {
  p <- ggplot(
    alzheimer_long[!is.na(bprs)],
    aes_string(x = "time", y = "bprs", color = var)
  ) +
    geom_smooth(method = "loess", se = FALSE) +
    labs(
      title = plot_titles[var],
      x = "Time (Years)",
      y = "BPRS Score"
    ) +
    theme_minimal() +
    scale_color_manual(
      name = "",
      values =mycolors
    ) +
    theme(legend.position = "bottom")


  if (var == "trial") {
    p <- p + guides(color = guide_legend(ncol = 10, byrow = TRUE))
  }
  plot_list[[var]] <- p
}

names(plot_list) <- subgroup_vars

trial_centers_plot <- plot_list[["trial"]]
plot_list[["trial"]] <- NULL
# Arrange the plots in a grid
plot_list[["simple_avg_plot"]] <- simple_avg_plot


# create ordered grob list with overall plot first
glist <- plot_list[c(
  "simple_avg_plot",
  setdiff(
    names(plot_list),
    "simple_avg_plot"
  )
)]
glist <- unname(glist)


ncol <- 2
n_other <- length(glist) - 1
nrows <- 1 + ceiling(n_other / ncol)
layout_mat <- matrix(NA_integer_, nrow = nrows, ncol = ncol)
layout_mat[1, ] <- 1L
idx <- 2L
for (r in 2:nrows) {
  for (c in 1:ncol) {
    if (idx <= length(glist)) {
      layout_mat[r, c] <- idx
      idx <- idx + 1L
    }
  }
}

## add figure 2 caption
bottom_caption <- grid::textGrob(
  "Figure 2: LOESS Smoothed BPRS Over Time by Subgroups.
  Overall average (top),  sex, education, job status, 
  living situation, and clinical center.",
  gp = grid::gpar(fontsize = 10),
  x = 0.5,
  hjust = 0.5
)

grid.arrange(
  grobs = glist,
  layout_matrix = layout_mat
  #bottom = bottom_caption
)
```

The exploratory analysis in Assignment 1 indicated (i) a clear increase in the mean BPRS trajectory over time, (ii) strong within-subject dependence (as reflected by the increasing semi-variogram with time lag), and (iii) substantial monotone dropout with missingness increasing markedly at later visits. The fitted linear mixed-effects model is in broad agreement with these empirical features: the fixed effect of time is strongly positive (\(\hat\beta_{\text{time}}=7.22\), \(p<2\times10^{-16}\)), supporting an increasing mean structure, while the inclusion of subject-specific random intercepts and slopes (\(\widehat{\mathrm{Var}}(b_{0i})=1.443\), \(\widehat{\mathrm{Var}}(b_{1i})=0.685\)) captures between-subject heterogeneity and induces correlation among repeated measurements within a patient. Moreover, a random-slope specification implies a marginal variance that can increase with follow-up time, which is consistent with the qualitative pattern suggested by the smoothed empirical variance function. while the inclusion of baseline covariates (sex, education, job status, living situation, age, BMI, ADL, and baseline CDR-SB) captures the group-level differences observed in the LOESS plots. 

Exact agreement between exploratory summaries and the fitted mixed model is not required, and moderate differences do not automatically indicate model inadequacy. Empirical variance functions and nonparametric LOESS curves are descriptive tools that are sensitive to local data density; in particular, variance estimates at later visits are computed from a progressively smaller and more selected subset of participants due to dropout, making them inherently less stable and potentially affected by selection. In contrast, the fitted mixed model imposes a parsimonious parametric structure on both the mean and covariance: it summarizes the mean trajectory using a linear (fixed-effect) time trend and models within-subject dependence through random effects, while typically assuming a constant residual variance (here \(\widehat{\sigma}^2 = 2.697\)) unless time-varying heteroscedasticity is explicitly introduced. Consequently, the model is not expected to reproduce the raw empirical variance at each visit or match the LOESS smooths exactly. The relevant diagnostic criterion is whether the model captures the dominant mean trend and main covariate effects, provides an adequate description of the within-subject correlation structure, and yields acceptable residual behaviour for the inferential goals of the analysis.

## Sensitivity analysis for the continuous outcome (BPRS)

### Sensitivity analysis for the continuous outcome (BPRS) under MNAR
Given (i) increasing missingness over time and (ii) evidence that missingness relates to observed severity (baseline BPRS vs missed visits), sensitivity analysis focuses on plausible MNAR departures where unobserved post-dropout BPRS differs systematically from MAR-based predictions. 

To assess the potential impact of MNAR departures on substantive conclusions, a pattern-mixture \(\delta\)-adjustment sensitivity analysis was performed. First, the primary likelihood-based MAR linear mixed model for BPRS was fitted and used to generate MAR-based predictions \(\widehat{Y}^{MAR}_{ij}\) for all subject–visit combinations. For post-dropout occasions (\(j>t_i\)) with missing BPRS, MNAR-adjusted values were constructed as
$$
Y^{mis}_{ij}(\delta)=\widehat{Y}^{MAR}_{ij}+\delta,
$$
where \(\delta\) represents a systematic shift in the unobserved post-dropout outcomes relative to MAR predictions. The primary mixed model was then refitted over a grid of clinically plausible \(\delta\) values (negative \(\delta\) representing worse unobserved psychiatric status than predicted under MAR, and positive \(\delta\) the opposite), and the stability of key fixed-effect conclusions (notably the time effect and the time-by-baseline severity interaction) was evaluated. This provides a transparent “robustness curve” describing how strong MNAR departures would need to be to materially alter inference, acknowledging that MAR is not empirically verifiable from observed data alone.

This sensitivity analysis is anchored to the observed missingness–severity relationship in this cohort and provides a transparent way to quantify how strongly conclusions would need to depend on unobserved BPRS values to materially change inference. The need for such an analysis follows directly from the fact that the MAR assumption is not empirically verifiable. 



If you paste the *exact variable names* you use for (a) the time index and (b) the baseline covariate set you decided to carry forward (the ones you consistently used in your BPRS model), I can tighten the notation so the missingness model and $\delta$-sensitivity section match your final model specification perfectly (same $X_i$, same centering, same time coding).


```{r mnar-delta-sensitivity, message=FALSE, warning=FALSE}
library(data.table)
library(lme4)
library(lmerTest)

setDT(alzheimer_long)
setorder(alzheimer_long, patid, time)

# If time is 1..7, convert to 0..6
# alzheimer_long[, time := time - 1L]

# ---------------------------
# 1) Fit the primary MAR model
# ---------------------------
mar_fit <- lmer(
  bprs ~ time + age + sex + bmi + job + adl + wzc + cdrsb0 + time:cdrsb0 + (time | patid),
  data = alzheimer_long,
  REML = FALSE
)

# ---------------------------
# 2) Identify dropout time t_i (last observed BPRS time)
#    and mark post-dropout missing occasions (time > t_i)
# ---------------------------
alzheimer_long[, R_bprs := as.integer(!is.na(bprs))]

ti_dt <- alzheimer_long[R_bprs == 1, .(t_i = max(time)), by = patid]
alzheimer_long <- merge(alzheimer_long, ti_dt, by = "patid", all.x = TRUE)

# Post-dropout missingness indicator (for monotone dropout this captures all missing)
alzheimer_long[, post_dropout_missing := (is.na(bprs) & time > t_i)]

# ---------------------------
# 3) Get MAR-based predictions for all rows (subject-specific BLUP predictions)
# ---------------------------
alzheimer_long[, bprs_hat_mar := predict(mar_fit, newdata = alzheimer_long, re.form = NULL)]

# ---------------------------
# 4) Delta-adjustment grid and refit model under each delta
# ---------------------------
delta_grid <- c(-10, -5, -2, 0, 2, 5, 10)   # choose values meaningful on BPRS scale

fit_under_delta <- function(delta) {
  dt <- copy(alzheimer_long)

  # Impute ONLY post-dropout missing values: Y_mis(delta) = Yhat_MAR + delta
  dt[post_dropout_missing == TRUE, bprs := bprs_hat_mar + delta]

  # Refit the same model to the delta-imputed dataset
  fit <- lmer(
    bprs ~ time + age + sex + bmi + job + adl + wzc + cdrsb0 + time:cdrsb0 + (time | patid),
    data = dt,
    REML = FALSE
  )

  # Extract key fixed effects (edit terms if you want more)
  cf <- summary(fit)$coefficients
  terms_keep <- c("time", "time:cdrsb0")
  out <- data.table(
    delta = delta,
    term = terms_keep,
    estimate = cf[terms_keep, "Estimate"],
    se = cf[terms_keep, "Std. Error"],
    p_value = cf[terms_keep, "Pr(>|t|)"]
  )
  out
}

sens_results <- rbindlist(lapply(delta_grid, fit_under_delta))

sens_results

```


```{r}
library(ggplot2)

ggplot(sens_results, aes(x = delta, y = estimate, group = term)) +
  geom_line() + geom_point() +
  facet_wrap(~ term, scales = "free_y") +
  labs(title = "MNAR sensitivity (delta-adjustment): fixed-effect estimates vs delta",
       x = expression(delta), y = "Estimate") +
  theme_minimal()
```


### MNAR sensitivity analysis (pattern-mixture \(\delta\)-adjustment)

A \(\delta\)-adjustment sensitivity analysis was used to assess how departures from MAR could affect the main mixed-model conclusions for BPRS. Post-dropout missing BPRS values were shifted relative to their MAR-based predictions by \(Y^{mis}_{ij}(\delta)=\widehat{Y}^{MAR}_{ij}+\delta\), and the primary mixed model was refitted over a grid of \(\delta\in\{-10,-5,-2,0,2,5,10\}\). Across this range, the substantive conclusions were highly robust. The estimated time effect remained strongly positive and highly significant for all \(\delta\) values, varying smoothly from 6.22 per year at \(\delta=-10\) to 8.22 per year at \(\delta=10\) (all \(p<10^{-16}\)), indicating that the conclusion of increasing BPRS over time does not depend on the MAR assumption. Similarly, the time-by-baseline severity interaction remained negative and statistically significant throughout, ranging from \(-0.0155\) to \(-0.0285\) (all \(p \le 6.5\times10^{-4}\)), implying that higher baseline CDR-SB is consistently associated with a slower increase in BPRS over time. Overall, even under sizable MNAR shifts in post-dropout outcomes, the direction and statistical significance of the key fixed effects were preserved, suggesting that the main longitudinal inferences are not sensitive to plausible MNAR departures in this dataset.







### Draft 2

### Missing data: assessment, primary (MAR) analysis, and MNAR sensitivity

Longitudinal follow-up in Alzheimer cohorts typically exhibits increasing missingness over time (e.g., missed visits, institutionalisation, illness progression, death). Because later visits are observed on a selected subset of patients, the missing-data strategy was specified *a priori* and integrated with the outcome models.

#### Notation

For subject $i=1,\dots,N$ at visit $j=0,\dots,J$, let $Y_{ij}$ denote the outcome of interest (e.g., continuous BPRS for the psychiatric symptom trajectory). Let

$$
R_{ij} =
\begin{cases}
1, & \text{if } Y_{ij} \text{ is observed},\
0, & \text{if } Y_{ij} \text{ is missing}.
\end{cases}
$$

Write $Y_i = (Y_i^{o}, Y_i^{m})$ for the observed and missing components of the full trajectory, and let $X_i$ denote baseline covariates (e.g., age, sex, education, baseline severity markers).



## 1) Missingness pattern and dropout characterisation (data-driven diagnostics)

The first step was to describe *how* missingness occurs in this dataset, since the pattern informs which modeling tools are appropriate and how sensitivity analyses should be designed.

### 1.1 Missingness pattern: complete, monotone, non-monotone

For each subject, the vector $R_i = (R_{i0},\dots,R_{iJ})$ was summarised as a binary “response string” (e.g., $1111000$). Subjects were classified as:

* **Complete** if $R_{ij}=1$ for all scheduled visits.

* **Monotone** if missingness begins and then persists. Equivalently, there exists a last observed time $t_i$ such that

$$
R_{ij}=1 \ \text{for } j \le t_i,
\qquad
R_{ij}=0 \ \text{for } j>t_i.
$$

* **Non-monotone** (intermittent) if there is at least one missing visit followed by a later observed visit, i.e., there exist $j<k$ such that $R_{ij}=0$ and $R_{ik}=1$.

### 1.2 Dropout pattern: complete, dropout, intermittent

A dropout time was defined as

$$
t_i = \max{ j : R_{ij}=1 }.
$$

Subjects were summarised by (i) whether they completed follow-up, (ii) the visit of last observation $t_i$, and (iii) whether missingness was intermittent (return after a missed visit). This description provides a clinically interpretable link between follow-up attrition and disease course (e.g., subjects with worse observed BPRS history may be more likely to drop out earlier).

### 1.3 Empirical evidence against MCAR (association with observed history)

Because MCAR implies that missingness is unrelated to observed outcomes, we assessed whether the probability of being observed depended on the observed history. A discrete-time model for observation at visit $j+1$ was used as a diagnostic:

$$
\Pr(R_{i,j+1}=1 \mid \mathcal{H}_{ij})
=
\operatorname{logit}^{-1}!\Big(
\alpha_0 + \alpha_1 Y_{ij} + \alpha_2^\top X_i + \alpha_3 j
\Big),
$$

where $\mathcal{H}_{ij}$ denotes the observed history up to visit $j$. A clear dependence on previously observed BPRS (or other observed markers) is consistent with missingness being related to observed data, which argues against MCAR and motivates methods valid under MAR as the primary analytic framework.



## 2) Primary missing-data assumption and why “MCAR/simple” methods were not used

### 2.1 Why not complete-case (CC) or LOCF

Two common “simple” approaches under an MCAR mindset are complete-case (CC) analysis and last observation carried forward (LOCF). In longitudinal Alzheimer follow-up, these can be problematic:

* **CC** uses only subjects with complete trajectories and can induce bias if completion is related to severity or prior outcomes; it also reduces efficiency by discarding partially observed profiles.

* **LOCF** implicitly assumes stability after the last observation and can distort the mean trajectory and variance structure, especially when symptoms tend to evolve over time.

Given the observed dropout/intermittency patterns and the clinical plausibility that missingness relates to observed history, the primary analysis proceeded under MAR using methods that retain all observed data without imputing ad hoc values.



## 3) MAR analysis strategies used for the study outcomes

Under MAR, the missingness mechanism can depend on observed history but not on unobserved outcomes after conditioning on observed data:

$$
f(R_i \mid X_i, Y_i^{o}, Y_i^{m})
=
f(R_i \mid X_i, Y_i^{o}).
$$

This assumption supports likelihood-based mixed models for subject-specific inference, and marginal models via GEE (with appropriate handling of incomplete data).

### 3.1 Continuous outcome (BPRS): direct likelihood via a linear mixed model

For the continuous BPRS trajectory, a linear mixed model was fitted by maximum likelihood (or REML for variance components where appropriate), using all available BPRS observations per subject. A general form is

$$
Y_{ij}
======

x_{ij}^\top \beta
+
z_{ij}^\top b_i
+
\varepsilon_{ij},
$$

with

$$
b_i \sim N(0, D),
\qquad
\varepsilon_{ij} \sim N(0,\sigma^2),
\qquad
b_i \perp \varepsilon_{ij}.
$$

Here, $x_{ij}$ contains fixed-effect terms (e.g., time and other covariates) and $z_{ij}$ defines the random-effects structure (e.g., random intercept and potentially random slope for time). The likelihood-based fit provides valid inference under MAR (given correct specification of the mean structure and random-effects distribution), while using partially observed trajectories efficiently.

### 3.2 Binary outcome: random-effects model (GLMM) under MAR

For the binary longitudinal endpoint, a subject-specific (conditional) model was fitted using a GLMM with a logit link:

$$
\Pr(Y_{ij}=1 \mid b_i, X_i)
=
\operatorname{logit}^{-1}!\Big(
x_{ij}^\top \beta + z_{ij}^\top b_i
\Big),
$$

with $b_i \sim N(0,D)$. This formulation captures within-subject correlation through random effects and yields conditional (subject-specific) interpretations of covariate effects.

### 3.3 Binary outcome: marginal model (GEE) under MAR

A marginal model was also fitted using GEE to obtain population-averaged effects:

$$
\Pr(Y_{ij}=1 \mid X_i)
=
\operatorname{logit}^{-1}\Big(
x_{ij}^\top \beta
\Big).
$$

Let $\mu_i(\beta)$ be the mean vector and $V_i$ a working covariance matrix. The estimating equation is

$$
\sum_{i=1}^N
D_i^\top
V_i^{-1}
\left(
Y_i - \mu_i(\beta)
\right)
=
0,
$$

where $D_i = \partial \mu_i(\beta)/\partial \beta^\top$. Robust (sandwich) standard errors were used to protect inference against misspecification of the working correlation.

### 3.4 Weighted GEE as an MAR approach when dropout is informative given observed history

When dropout depends on observed history, inverse probability weighting provides a principled MAR correction for marginal inference. Let $\pi_{ij} = \Pr(R_{ij}=1 \mid \mathcal{H}_{i,j-1})$ be estimated from an observation model (such as the logistic diagnostic model above). Define weights

$$
w_{ij} = \frac{1}{\pi_{ij}}.
$$

Weighted GEE solves

$$
\sum_{i=1}^N
D_i^\top
V_i^{-1}
W_i
\left(
Y_i - \mu_i(\beta)
\right)
=
0,
$$

where $W_i$ is typically diagonal with entries $w_{ij}$ for observed components. Weight behaviour (e.g., presence of extreme weights) was monitored as a practical diagnostic of whether observed history adequately explains missingness.

### 3.5 Multiple imputation (MI) as an MAR alternative (if included)

As an MAR-based alternative, multiple imputation can be used to create $M$ completed datasets by drawing $Y^{m}$ from an imputation model conditional on $(Y^{o},X)$. Each dataset is analysed with the chosen outcome model, and estimates are combined using Rubin’s rules. Inference targets and model specifications match those of the primary analyses, while MI propagates uncertainty due to missingness through between-imputation variability.

---

## 4) MNAR: sensitivity analysis anchored on BPRS (continuous outcome)

Because MNAR implies dependence of missingness on unobserved outcomes,

$$
f(R_i \mid X_i, Y_i^{o}, Y_i^{m}) \ \text{depends on } Y_i^{m},
$$

it cannot be empirically verified from the observed data alone. Therefore, robustness was assessed via a structured sensitivity analysis using the BPRS trajectory, reflecting the clinical concern that patients may become unobserved because their unmeasured symptom burden is systematically worse than expected under MAR.

### 4.1 Pattern-mixture $\delta$-adjustment after dropout

Let $\widehat{Y}^{MAR}_{ij}$ denote the MAR-based predicted value (from the fitted linear mixed model). For post-dropout visits $j>t_i$, define an MNAR-adjusted missing outcome as

$$
Y_{ij}^{m}(\delta)
=
\widehat{Y}^{MAR}_{ij}
+
\delta,
\qquad
j>t_i.
$$

Here, $\delta$ encodes the degree and direction of MNAR departure:

* $\delta=0$ corresponds to MAR.
* $\delta>0$ represents systematically higher (worse) unobserved BPRS after dropout than MAR would predict.
* $\delta<0$ represents systematically lower unobserved BPRS.

Key estimands (e.g., time trend and covariate effects) were recomputed across a pre-specified grid of $\delta$ values to evaluate how conclusions change as missingness becomes increasingly non-ignorable. This approach provides an interpretable robustness summary in the Alzheimer context: it quantifies the magnitude of unobserved symptom shift required to materially alter inference.

### 4.2 Interpretation

Sensitivity analysis results were reported as ranges (or profiles) of the main parameters of interest across $\delta$, emphasizing whether substantive conclusions are stable under clinically plausible MNAR deviations.

---

If you paste the exact fixed-effect structure you used for time (e.g., linear time, quadratic time, or categorical visit indicators) and which binary outcome you modelled, I can rewrite the model equations *verbatim* to match your final formulas (same covariate symbols, same link functions, same random-effects structure).



## Results


```{r missingness-taxonomy-code, echo=TRUE}
#| label: q1-missingness-taxonomy-code
#| message: false
#| warning: false

library(data.table)
library(ggplot2)

# ------------------------------------------------------------
# 1) Taxonomy helpers: complete / monotone / non-monotone
#    and complete / dropout / intermittent
# ------------------------------------------------------------

classify_missingness <- function(dt, id, cols, time = NULL) {
  stopifnot(is.data.table(dt))
  if (is.null(time)) time <- seq_along(cols) - 1L

  # enforce time ordering
  ord  <- order(time)
  cols <- cols[ord]
  time <- time[ord]

  # R_ij = 1 if observed, 0 if missing
  R_dt <- dt[, lapply(.SD, function(x) as.integer(!is.na(x))), .SDcols = cols]
  Rmat <- as.matrix(R_dt)

  # complete cases
  is_complete <- rowSums(1L - Rmat) == 0L

  # monotone missingness (dropout-type): once missing then always missing
  is_monotone <- apply(Rmat, 1, function(r) {
    if (all(r == 1L)) return(TRUE)                 # complete is monotone, but we separate later
    last1 <- if (any(r == 1L)) max(which(r == 1L)) else 0L
    all(r[(last1 + 1L):length(r)] == 0L)
  })

  # non-monotone: intermittent gaps exist
  is_nonmon <- !is_monotone

  # dropout time T_i = last observed visit time
  last_obs_idx <- apply(Rmat, 1, function(r) if (any(r == 1L)) max(which(r == 1L)) else NA_integer_)
  T_i <- ifelse(is.na(last_obs_idx), NA_integer_, time[last_obs_idx])

  pattern_missingness <- fifelse(is_complete, "complete",
                                fifelse(is_monotone, "monotone", "non-monotone"))
  pattern_dropout <- fifelse(is_complete, "complete",
                             fifelse(is_monotone, "dropout", "intermittent"))

  pattern_string <- apply(Rmat, 1, paste0, collapse = "")

  out <- dt[, .(patid = get(id))]
  out[, `:=`(
    missingness_pattern = pattern_missingness,
    dropout_pattern     = pattern_dropout,
    last_observed_time  = T_i,
    n_missing           = rowSums(1L - Rmat),
    pattern             = pattern_string
  )]
  out
}

summarise_missing_by_time <- function(dt, id, cols, time = NULL) {
  stopifnot(is.data.table(dt))
  if (is.null(time)) time <- seq_along(cols) - 1L

  ord  <- order(time)
  cols <- cols[ord]
  time <- time[ord]

  long <- melt(
    dt[, c(id, cols), with = FALSE],
    id.vars = id,
    variable.name = "visit",
    value.name = "y",
    variable.factor = FALSE
  )
  long[, time := time[match(visit, cols)]]
  long[, observed := !is.na(y)]

  long[, .(
    n_patients  = uniqueN(get(id)),
    n_obs       = sum(observed),
    pct_missing = 100 * (uniqueN(get(id)) - sum(observed)) / uniqueN(get(id))
  ), by = time][order(time)]
}

# ------------------------------------------------------------
# 2) Apply to the continuous outcome (BPRS)
# ------------------------------------------------------------

bprs_cols <- paste0("bprs", 0:6)
tax_bprs  <- classify_missingness(alzheimer_data, id = "patid", cols = bprs_cols, time = 0:6)
miss_t_bprs <- summarise_missing_by_time(alzheimer_data, id = "patid", cols = bprs_cols, time = 0:6)

# summaries for reporting
tab_missingness_bprs <- tax_bprs[, .(n = .N), by = missingness_pattern][order(missingness_pattern)]
tab_missingness_bprs[, pct := 100 * n / sum(n)]

tab_dropout_bprs <- tax_bprs[, .(n = .N), by = dropout_pattern][order(dropout_pattern)]
tab_dropout_bprs[, pct := 100 * n / sum(n)]

top_patterns_bprs <- tax_bprs[, .(n = .N), by = pattern][order(-n)][1:10]

# ------------------------------------------------------------
# 3) Plots: missingness by time + pattern heatmap (optional)
# ------------------------------------------------------------

p_miss_time_bprs <- ggplot(miss_t_bprs, aes(x = time, y = pct_missing)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(x = "Time (years)", y = "Percent missing", title = "Missingness over time (BPRS)") +
  theme_minimal()

# Heatmap of individual patterns (use a sample if too many patients)
set.seed(1)
tax_sample <- tax_bprs[sample(.N, min(.N, 200L))]

# Expand the pattern string into columns for plotting
pat_mat <- t(vapply(strsplit(tax_sample$pattern, ""), as.integer, integer(7)))
colnames(pat_mat) <- paste0("t", 0:6)

heat_dt <- data.table(patid = tax_sample$patid, pat_mat)
heat_long <- melt(heat_dt, id.vars = "patid", variable.name = "time", value.name = "R")
heat_long[, time := as.integer(sub("^t", "", time))]

p_heat_bprs <- ggplot(heat_long, aes(x = time, y = factor(patid), fill = factor(R))) +
  geom_tile() +
  labs(x = "Time (years)", y = "Patient (sample)", title = "Observation indicator R_ij (BPRS)") +
  theme_minimal() +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank()) +
  guides(fill = "none")

# print(p_miss_time_bprs)
# print(p_heat_bprs)

# ------------------------------------------------------------
# 4) Objects that map to SEM / PMM / SPM frameworks
#    (these are the building blocks you will reuse later)
# ------------------------------------------------------------

# SEM-style: long data with R_ij and observed history (lagged outcome)
# (works off your existing long dataset)
alz_sem_bprs <- copy(alzheimer_long)
alz_sem_bprs[, R_bprs := as.integer(!is.na(bprs))]
setorder(alz_sem_bprs, patid, time)
alz_sem_bprs[, bprs_lag1 := shift(bprs, 1L, type = "lag"), by = patid]
alz_sem_bprs[, dbprs_lag1 := bprs_lag1 - shift(bprs, 2L, type = "lag"), by = patid]

# PMM-style: dropout groups via T_i (pattern-mixture stratification variable)
alz_pmm_bprs <- merge(alzheimer_long, tax_bprs, by = "patid", all.x = TRUE)
alz_pmm_bprs[, dropout_group := factor(last_observed_time, levels = 0:6)]

# SPM-style: for shared-parameter ideas you typically need the same ingredients:
# - longitudinal outcome (bprs)
# - missingness indicator (R_bprs)
# - shared random effects (later in modelling, not in preprocessing)
# So the key dataset is alz_sem_bprs (contains both bprs and R_bprs).

```
















